---
title: "Fish provisioning"
subtitle: "EU marine ecosystem service accounts, 2008 to 2023"
format:
  dashboard:
    theme:
      - cosmo
      - brand
    scrolling: true
    orientation: rows
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| output: false

library(tidyverse)
library(sf)
library(gt)

sf_use_s2(FALSE)

reference_year <- 2023
base_year <- 2010

fish_provisioning <- readRDS("data/outputs/results/fish_provisioning.RDS")
country_names <- readr::read_csv(
  "data/stecf/country_codes.csv",
  show_col_types = FALSE
) |>
  transmute(country_code = `alpha-3`, country = name) |>
  mutate(
    country = recode(country, `Netherlands, Kingdom of the` = "Netherlands")
  )
macro <- st_read(
  "data/outputs/gis/fish_provisioning.gpkg",
  layer = "sea_regions_macro",
  quiet = TRUE
)

sea_order <- c(
  "Baltic Sea", "Black Sea", "Mediterranean Sea", "NEAO", "ROW", "unknown"
)
nested_seas <- c(
  "Baltic Sea", "Black Sea", "Mediterranean Sea", "NEAO", "ROW"
)

sut_value_cols <- c(
  "fish_benthopelagic", "fish_demersal", "fish_pelagic",
  "shellfish_benthic", "shellfish_benthopelagic",
  "shellfish_demersal", "shellfish_pelagic",
  "other_not_identified"
)

sut_type_labels <- c(
  fish_benthopelagic = "benthopelagic",
  fish_demersal = "demersal",
  fish_pelagic = "pelagic",
  shellfish_benthic = "benthic",
  shellfish_benthopelagic = "benthopelagic",
  shellfish_demersal = "demersal",
  shellfish_pelagic = "pelagic",
  other_not_identified = "Not identified",
  Total = "Total"
)

prepare_sut_types <- function(data) {
  data |>
    mutate(
      species_group = case_when(
        species_group == "fish" ~ "fish",
        species_group == "shellfish" ~ "shellfish",
        .default = "other"
      ),
      fish_type = case_when(
        species_group == "other" ~ "not_identified",
        is.na(fish_type) ~ "not_identified",
        .default = fish_type
      ),
      species_group = factor(species_group, c("fish", "shellfish", "other")),
      fish_type = factor(
        fish_type,
        c("benthic", "benthopelagic", "demersal", "pelagic", "not_identified")
      )
    )
}

nested_sut <- function(data, row_var, value_var) {
  wide <- data |>
    prepare_sut_types() |>
    summarise(
      value = sum(.data[[value_var]], na.rm = TRUE),
      .by = all_of(c(row_var, "species_group", "fish_type"))
    ) |>
    pivot_wider(
      names_from = c(species_group, fish_type),
      values_from = value,
      names_sep = "_",
      values_fill = 0,
      names_expand = TRUE
    )

  missing_cols <- setdiff(sut_value_cols, names(wide))
  for (col in missing_cols) {
    wide[[col]] <- 0
  }

  wide |>
    mutate(Total = rowSums(across(all_of(sut_value_cols)), na.rm = TRUE)) |>
    filter(Total != 0) |>
    select(all_of(row_var), all_of(sut_value_cols), Total) |>
    arrange(across(all_of(row_var)))
}

add_total_row <- function(tbl, row_var) {
  totals <- tbl |>
    summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE))) |>
    mutate(!!row_var := "Total", .before = 1)
  bind_rows(tbl, totals)
}

sea_sut <- function(year, sea, side, value_var) {
  slice <- fish_provisioning |>
    filter(year == .env$year, Sea_reg == .env$sea)
  if (side == "supply") {
    nested_sut(slice, "sub_reg_clean", value_var) |>
      add_total_row("sub_reg_clean") |>
      rename(EEZ = sub_reg_clean)
  } else {
    slice |>
      left_join(country_names, by = "country_code") |>
      mutate(country = coalesce(country, country_code)) |>
      nested_sut("country", value_var) |>
      add_total_row("country") |>
      rename(Country = country)
  }
}

sut_gt <- function(tbl, title) {
  row_col <- names(tbl)[1]
  tbl |>
    gt(rowname_col = row_col) |>
    cols_label(.list = as.list(sut_type_labels)) |>
    tab_spanner("Fish", c(fish_benthopelagic, fish_demersal, fish_pelagic)) |>
    tab_spanner(
      "Shellfish",
      c(
        shellfish_benthic, shellfish_benthopelagic,
        shellfish_demersal, shellfish_pelagic
      )
    ) |>
    tab_spanner("Others", other_not_identified) |>
    fmt_number(decimals = 2) |>
    tab_header(title = title) |>
    tab_stubhead(label = row_col)
}

print_sea_year <- function(y, sea) {
  has_rows <- any(
    fish_provisioning$year == y & fish_provisioning$Sea_reg == sea
  )
  if (!has_rows) {
    return(invisible())
  }
  print(sut_gt(
    sea_sut(y, sea, "supply", "landed_tonnes"),
    paste0(sea, " supply, physical (t), ", y)
  ))
  print(sut_gt(
    sea_sut(y, sea, "use", "landed_tonnes"),
    paste0(sea, " use, physical (t), ", y)
  ))
  print(sut_gt(
    sea_sut(y, sea, "supply", "revenue_thousand_eur_constant"),
    paste0(sea, " supply, thousand ", base_year, " euro, ", y)
  ))
  print(sut_gt(
    sea_sut(y, sea, "use", "revenue_thousand_eur_constant"),
    paste0(sea, " use, thousand ", base_year, " euro, ", y)
  ))
}

by_sea_year <- fish_provisioning |>
  filter(Sea_reg != "unknown") |>
  summarise(
    landed_tonnes = sum(landed_tonnes),
    revenue_thousand_eur = sum(revenue_thousand_eur),
    revenue_thousand_eur_constant = sum(revenue_thousand_eur_constant),
    .by = c(year, Sea_reg)
  ) |>
  mutate(
    price_eur_per_kg_constant = if_else(
      landed_tonnes > 0,
      revenue_thousand_eur_constant / landed_tonnes,
      NA_real_
    ),
    Sea_reg = factor(Sea_reg, levels = sea_order)
  ) |>
  arrange(Sea_reg, year)

panel_years <- sort(unique(by_sea_year$year), decreasing = TRUE)

by_sea_2023 <- by_sea_year |>
  filter(year == reference_year) |>
  arrange(Sea_reg)

totals_2023 <- by_sea_2023 |>
  summarise(
    landed_tonnes = sum(landed_tonnes),
    revenue_thousand_eur = sum(revenue_thousand_eur),
    revenue_thousand_eur_constant = sum(revenue_thousand_eur_constant)
  )

row_share_2023 <- 100 *
  by_sea_2023$landed_tonnes[by_sea_2023$Sea_reg == "ROW"] /
  totals_2023$landed_tonnes

macro_2023 <- macro |>
  mutate(
    landed_tonnes = tonnes_2023,
    revenue_thousand_eur_constant = rent_const_k_eur_2023,
    price_eur_per_kg_constant = price_const_eur_kg_2023,
    Sea_label = recode(
      Sea_name,
      NEAO = "North East Atlantic",
      MED = "Mediterranean Sea",
      Black = "Black Sea",
      Baltic = "Baltic Sea"
    )
  )

choropleth <- function(data, fill, legend) {
  ggplot(data) +
    geom_sf(aes(fill = {{ fill }}), colour = "white", linewidth = 0.35) +
    scale_fill_viridis_c(
      option = "C",
      labels = scales::label_comma(),
      name = legend,
      na.value = "grey90"
    ) +
    coord_sf(expand = FALSE) +
    theme_void() +
    theme(
      legend.position = "right",
      plot.caption = element_text(hjust = 0, size = 10, colour = "grey30"),
      plot.margin = margin(4, 8, 4, 8)
    )
}

fmt_tonnes <- function(x) {
  paste0(format(round(x), big.mark = ",", scientific = FALSE), " t")
}

fmt_million_eur <- function(x) {
  paste0(
    format(round(x / 1000), big.mark = ",", scientific = FALSE),
    " million"
  )
}
```

# Overview

## Row {height="16%"}

```{r}
#| content: valuebox
#| title: "2023 landings"
#| icon: globe-europe-africa
#| color: primary

list(value = fmt_tonnes(totals_2023$landed_tonnes))
```

```{r}
#| content: valuebox
#| title: "2023 resource rent, 2010 euro"
#| icon: currency-euro
#| color: secondary

list(value = fmt_million_eur(totals_2023$revenue_thousand_eur_constant))
```

```{r}
#| content: valuebox
#| title: "ROW share of 2023 tonnage"
#| icon: globe
#| color: warning

list(value = paste0(round(row_share_2023), "%"))
```

## Row {height="12%"}

::: {.card}
Nested supply and use tables follow methods Tables 10 to 13. Supply rows are FAO/GFCM fishing
areas, labelled EEZ as the spatial unit of supply; they are not legal exclusive economic zones.
Use rows are the Member State of landing. Columns are species group crossed with fish type.
Supply Total equals use Total. Nominal euro tables are in the Excel workbook; the dashboard
shows physical tonnes and 2010 euro.
:::

::: {.card}
**Monetary values are 2010 euro.** Resource rent is deflated with Eurostat HICP for EU-27.
The allocation itself stays in nominal euro. ROW is rest of world.
:::

# Supply and use

## Row {height="88%"}

::: {.card title="Nested SUT by sea (physical and 2010 euro)"}
::: {.panel-tabset}

```{r}
#| label: sut-tabs
#| results: asis

for (sea in nested_seas) {
  cat("\n### ", sea, "\n\n", sep = "")
  cat("::: {.panel-tabset}\n\n")
  for (y in panel_years) {
    cat("\n#### ", y, "\n\n", sep = "")
    print_sea_year(y, sea)
    cat("\n")
  }
  cat("\n:::\n\n")
}
```

:::
:::

# Maps

## Row {height="12%"}

::: {.card}
Choropleths show the four Marine Strategy Framework Directive regions only. There is no legal-EEZ
layer in the published GeoPackage. Supply tables label FAO/GFCM fishing areas as EEZ; use tables
are Member State of landing. ROW was
`r round(row_share_2023)` percent of 2023 tonnage and falls outside the European
map. Value and price are 2010 euro.
:::

## Row {.tabset height="88%"}

### Landings (tonnes)

```{r}
#| fig-width: 10
#| fig-height: 7

choropleth(macro_2023, landed_tonnes, "Tonnes") +
  labs(caption = "2023 live weight. ROW is not shown.")
```

### Value (thousand 2010 €)

```{r}
#| fig-width: 10
#| fig-height: 7

choropleth(macro_2023, revenue_thousand_eur_constant, "Thousand\n2010 €") +
  labs(caption = "2023 resource rent, 2010 euro. ROW is not shown.")
```

### Price (2010 €/kg)

```{r}
#| fig-width: 10
#| fig-height: 7

choropleth(macro_2023, price_eur_per_kg_constant, "2010 €/kg") +
  labs(caption = "2023 quantity-weighted resource rent per kilogram, 2010 euro. ROW is not shown.")
```

# Trends

## Row

```{r}
#| title: "Landings by sea region"
#| fig-width: 11
#| fig-height: 4.5

by_sea_year |>
  ggplot(aes(year, landed_tonnes, colour = Sea_reg)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.6) +
  scale_y_continuous(labels = scales::label_comma()) +
  scale_x_continuous(breaks = seq(2008, 2023, by = 2)) +
  scale_colour_brewer(palette = "Dark2", name = NULL) +
  labs(x = NULL, y = "Tonnes") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Row

```{r}
#| title: "Resource rent by sea region (2010 euro)"
#| fig-width: 11
#| fig-height: 4.5

by_sea_year |>
  ggplot(aes(year, revenue_thousand_eur_constant, colour = Sea_reg)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.6) +
  scale_y_continuous(labels = scales::label_comma()) +
  scale_x_continuous(breaks = seq(2008, 2023, by = 2)) +
  scale_colour_brewer(palette = "Dark2", name = NULL) +
  labs(
    x = NULL,
    y = "Thousand 2010 €",
    caption = "Constant 2010 euro, Eurostat HICP for EU-27."
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```
