---
title: "Fish provisioning"
subtitle: "EU marine ecosystem service accounts, 2008 to 2023"
format:
  dashboard:
    theme:
      - cosmo
      - brand
    scrolling: true
    orientation: rows
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| output: false

library(tidyverse)
library(sf)
library(gt)

sf_use_s2(FALSE)

reference_year <- 2023
base_year <- 2008

fish_provisioning <- readRDS("data/outputs/results/fish_provisioning.RDS")
macro <- st_read(
  "data/outputs/gis/fish_provisioning.gpkg",
  layer = "sea_regions_macro",
  quiet = TRUE
)

sea_order <- c(
  "NEAO", "Baltic Sea", "Mediterranean Sea", "Black Sea", "ROW"
)

by_sea_year <- fish_provisioning |>
  filter(Sea_reg != "unknown") |>
  summarise(
    landed_tonnes = sum(landed_tonnes),
    revenue_thousand_eur = sum(revenue_thousand_eur),
    revenue_thousand_eur_2008 = sum(revenue_thousand_eur_2008),
    .by = c(year, Sea_reg)
  ) |>
  mutate(
    price_eur_per_kg_2008 = if_else(
      landed_tonnes > 0,
      revenue_thousand_eur_2008 / landed_tonnes,
      NA_real_
    ),
    Sea_reg = factor(Sea_reg, levels = sea_order)
  ) |>
  arrange(Sea_reg, year) |>
  mutate(
    real_change_pct = 100 *
      (revenue_thousand_eur_2008 / lag(revenue_thousand_eur_2008) - 1),
    .by = Sea_reg
  )

panel_years <- sort(unique(by_sea_year$year), decreasing = TRUE)

by_sea_2023 <- by_sea_year |>
  filter(year == reference_year) |>
  arrange(Sea_reg)

totals_2023 <- by_sea_2023 |>
  summarise(
    landed_tonnes = sum(landed_tonnes),
    revenue_thousand_eur = sum(revenue_thousand_eur),
    revenue_thousand_eur_2008 = sum(revenue_thousand_eur_2008)
  ) |>
  mutate(
    price_eur_per_kg_2008 = revenue_thousand_eur_2008 / landed_tonnes
  )

row_share_2023 <- 100 *
  by_sea_2023$landed_tonnes[by_sea_2023$Sea_reg == "ROW"] /
  totals_2023$landed_tonnes

macro_2023 <- macro |>
  mutate(
    landed_tonnes = tonnes_2023,
    revenue_thousand_eur_2008 = rent08_k_eur_2023,
    price_eur_per_kg_2008 = price08_eur_kg_2023,
    Sea_label = recode(
      Sea_name,
      NEAO = "North East Atlantic",
      MED = "Mediterranean Sea",
      Black = "Black Sea",
      Baltic = "Baltic Sea"
    )
  )

supply_gt <- function(y) {
  by_sea <- by_sea_year |>
    filter(year == y) |>
    arrange(Sea_reg)
  totals <- by_sea |>
    summarise(
      landed_tonnes = sum(landed_tonnes),
      revenue_thousand_eur = sum(revenue_thousand_eur),
      revenue_thousand_eur_2008 = sum(revenue_thousand_eur_2008)
    ) |>
    mutate(
      price_eur_per_kg_2008 = revenue_thousand_eur_2008 / landed_tonnes,
      prev_total = sum(
        by_sea_year$revenue_thousand_eur_2008[by_sea_year$year == y - 1]
      ),
      real_change_pct = if_else(
        prev_total > 0,
        100 * (revenue_thousand_eur_2008 / prev_total - 1),
        NA_real_
      )
    )
  bind_rows(
    by_sea |>
      transmute(
        Sea_reg = as.character(Sea_reg),
        landed_tonnes,
        revenue_thousand_eur,
        revenue_thousand_eur_2008,
        price_eur_per_kg_2008,
        real_change_pct
      ),
    totals |>
      transmute(
        Sea_reg = "Total",
        landed_tonnes,
        revenue_thousand_eur,
        revenue_thousand_eur_2008,
        price_eur_per_kg_2008,
        real_change_pct
      )
  ) |>
    gt(rowname_col = "Sea_reg") |>
    cols_label(
      landed_tonnes = "Landings (tonnes)",
      revenue_thousand_eur = "Resource rent (thousand nominal €)",
      revenue_thousand_eur_2008 = "Resource rent (thousand 2008 €)",
      price_eur_per_kg_2008 = "Price (2008 €/kg)",
      real_change_pct = "Real change (%)"
    ) |>
    fmt_number(landed_tonnes, decimals = 0) |>
    fmt_number(revenue_thousand_eur, decimals = 0) |>
    fmt_number(revenue_thousand_eur_2008, decimals = 0) |>
    fmt_number(price_eur_per_kg_2008, decimals = 2) |>
    fmt_number(real_change_pct, decimals = 1) |>
    tab_source_note(
      "Real change is the year-on-year change in 2008 euro. Blank in 2008. 2024 is excluded because expenditure was not reported."
    )
}

choropleth <- function(data, fill, legend) {
  ggplot(data) +
    geom_sf(aes(fill = {{ fill }}), colour = "white", linewidth = 0.35) +
    scale_fill_viridis_c(
      option = "C",
      labels = scales::label_comma(),
      name = legend,
      na.value = "grey90"
    ) +
    coord_sf(expand = FALSE) +
    theme_void() +
    theme(
      legend.position = "right",
      plot.caption = element_text(hjust = 0, size = 10, colour = "grey30"),
      plot.margin = margin(4, 8, 4, 8)
    )
}

fmt_tonnes <- function(x) {
  paste0(format(round(x), big.mark = ",", scientific = FALSE), " t")
}

fmt_million_eur <- function(x) {
  paste0(
    format(round(x / 1000), big.mark = ",", scientific = FALSE),
    " million"
  )
}
```

# Overview

## Row {height="16%"}

```{r}
#| content: valuebox
#| title: "2023 landings"
#| icon: globe-europe-africa
#| color: primary

list(value = fmt_tonnes(totals_2023$landed_tonnes))
```

```{r}
#| content: valuebox
#| title: "2023 resource rent, 2008 euro"
#| icon: currency-euro
#| color: secondary

list(value = fmt_million_eur(totals_2023$revenue_thousand_eur_2008))
```

```{r}
#| content: valuebox
#| title: "ROW share of 2023 tonnage"
#| icon: globe
#| color: warning

list(value = paste0(round(row_share_2023), "%"))
```

## Row {height="62%"}

::: {.card title="Supply by sea region (2008 euro)"}
::: {.panel-tabset}

```{r}
#| label: supply-tabs
#| results: asis

for (y in panel_years) {
  cat("\n### ", y, "\n\n", sep = "")
  print(supply_gt(y))
  cat("\n")
}
```

:::
:::

## Row {height="22%"}

::: {.card}
**Rest of world is off the map.** ROW was `r round(row_share_2023)` percent of 2023
tonnage. Those landings were taken by EU fleets outside the four pan-European sea
regions and cannot appear on a European choropleth. They remain in the tables and
in the time series.
:::

::: {.card}
**Monetary values are 2008 euro.** Resource rent is deflated with Eurostat HICP for
EU-27. The allocation itself stays in nominal euro. Real change is the year-on-year
change in 2008 euro. The base year is the first year of the panel and can be switched
in one line if the client prefers 2023 constant prices.
:::

# Maps

## Row {height="12%"}

::: {.card}
Choropleths show the four Marine Strategy Framework Directive regions only. ROW was
`r round(row_share_2023)` percent of 2023 tonnage and falls outside the European
map. Value and price are 2008 euro.
:::

## Row {.tabset height="88%"}

### Landings (tonnes)

```{r}
#| fig-width: 10
#| fig-height: 7

choropleth(macro_2023, landed_tonnes, "Tonnes") +
  labs(caption = "2023 live weight. ROW is not shown.")
```

### Value (thousand 2008 €)

```{r}
#| fig-width: 10
#| fig-height: 7

choropleth(macro_2023, revenue_thousand_eur_2008, "Thousand\n2008 €") +
  labs(caption = "2023 resource rent, 2008 euro. ROW is not shown.")
```

### Price (2008 €/kg)

```{r}
#| fig-width: 10
#| fig-height: 7

choropleth(macro_2023, price_eur_per_kg_2008, "2008 €/kg") +
  labs(caption = "2023 quantity-weighted resource rent per kilogram, 2008 euro. ROW is not shown.")
```

# Trends

## Row

```{r}
#| title: "Landings by sea region"
#| fig-width: 11
#| fig-height: 4.5

by_sea_year |>
  ggplot(aes(year, landed_tonnes, colour = Sea_reg)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.6) +
  scale_y_continuous(labels = scales::label_comma()) +
  scale_x_continuous(breaks = seq(2008, 2023, by = 2)) +
  scale_colour_brewer(palette = "Dark2", name = NULL) +
  labs(x = NULL, y = "Tonnes") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Row

```{r}
#| title: "Resource rent by sea region (2008 euro)"
#| fig-width: 11
#| fig-height: 4.5

by_sea_year |>
  ggplot(aes(year, revenue_thousand_eur_2008, colour = Sea_reg)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.6) +
  scale_y_continuous(labels = scales::label_comma()) +
  scale_x_continuous(breaks = seq(2008, 2023, by = 2)) +
  scale_colour_brewer(palette = "Dark2", name = NULL) +
  labs(
    x = NULL,
    y = "Thousand 2008 €",
    caption = "Constant 2008 euro, Eurostat HICP for EU-27. Year-on-year change is real change in the service."
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```
