---
title: "STECF Fleet Segment Data"
subtitle: "From one large workbook to two analysis-ready datasets"
format: html
toc: true
toc-depth: 3
execute:
  warning: false
  message: false
---

# Why this page exists

The fish provisioning accounts are built from a single Excel workbook published by the Scientific,
Technical and Economic Committee for Fisheries (STECF). That workbook is 285 MB, which is too large
to keep under version control and slow to open on most machines. This page records what the
workbook contains, how we extracted the parts we need, and what the extraction is safe to be used
for.

The conversion is done once by `data/stecf/stecf_to_rds.R`. Its two outputs are committed to the
repository; the workbook itself is listed in `.gitignore` and stays on local machines only. Anyone
who needs to regenerate the RDS files can download the workbook again and rerun the script.

# The source workbook

The file is `STECF_25 07_EU Fleet Economic and Transversal data_fleet segment level.xlsx`, released
under the 2025 Fleet Economic data call. It holds twelve sheets:

| Sheet | Content |
|---|---|
| `NOTES` | Coverage, confidentiality and definitions |
| `FS data` | Economic and transversal variables at fleet segment level |
| `pivot fs data` | Excel pivot of the above |
| `FS_effort by sub-region ` | Effort by fleet segment and FAO sub-region |
| `pivots_effort by sub-region` | Excel pivot of the above |
| `FS_landings FAO 2018-2023` | Landings by species and sub-region, 2018 to 2024 |
| `pivot_landings` | Excel pivot of the above |
| `FS_landings FAO 2013-2017` | Landings by species and sub-region |
| `FS_landings FAO 2008-2012` | Landings by species and sub-region |
| `additional info` | Country codes, gear codes, inflation and interest rates |
| `codes_species` | FAO ASFIS code, scientific name, common name |
| `codes__regions_sub-regions` | Sub-region code, FAO area, supra-region |

We extract four of them: the three landings sheets and `FS data`. The pivot sheets are Excel
artefacts of the same numbers. The effort sheets are not used by the fish provisioning method. The
three code sheets duplicate lookups we already curate in `data/lookup`, so we leave them where they
are for now.

::: {.callout-important}
## The sheet names understate their content

`FS_landings FAO 2018-2023` contains 2024 as well, and `FS data` runs to 2024. The output files are
therefore named for the years they actually hold, 2008 to 2024, rather than for the sheet labels.
:::

# What the script does

`data/stecf/stecf_to_rds.R` reads the four sheets, binds the three landings sheets into one table,
and writes two RDS files into `data/stecf/`. Run it from the project root:

```bash
Rscript data/stecf/stecf_to_rds.R
```

It takes about 35 seconds and turns 285 MB of Excel into 18 MB of RDS.

Three design decisions are worth recording.

**Column types are declared, not guessed.** Left to itself, `readxl` infers `cluster_name` as
logical in the 2018-2023 sheet, where that column happens to be empty, and as text in the other
two; binding the three together would then fail. The script declares all 24 types by name and
compares them against the actual header before reading. A future data call that renames or reorders
a column stops the script with an explicit message rather than producing a quietly wrong table.

**The script converts, it does not clean.** No rows are dropped and no values are recoded. Rows
where `value` is missing are kept as they appear in the workbook, because a missing value is
meaningful: it usually indicates that a Member State did not report that variable, or that the
figure was suppressed for confidentiality. Deciding what to do with them belongs to the analysis,
not to the format conversion.

**Each row records the sheet it came from.** A `source_sheet` column is added so that any question
about provenance can be answered from the dataset itself.

# The two datasets

```{r}
#| label: load

library(dplyr)
library(tidyr)
library(gt)

fs_landings <- readRDS("data/stecf/FS_landings_FAO_2008-2024.RDS")
fs_data <- readRDS("data/stecf/FS_data_2008-2024.RDS")
```

## `FS_landings_FAO_2008-2024.RDS`

One row per fleet segment, year, species and FAO sub-region. Every observation appears twice, once
as a live weight in kilograms and once as a value in euro.

```{r}
#| label: landings-shape

fs_landings |>
  summarise(
    rows = n(),
    years = paste(min(year), max(year), sep = " to "),
    countries = n_distinct(country_code),
    species = n_distinct(species_code),
    sub_regions = n_distinct(sub_reg),
    .by = source_sheet
  ) |>
  gt() |>
  fmt_number(rows, decimals = 0)
```

```{r}
#| label: landings-variables

fs_landings |>
  count(variable_group, variable_name, variable_code, unit) |>
  gt() |>
  fmt_number(n, decimals = 0)
```

## `FS_data_2008-2024.RDS`

One row per fleet segment, year and economic or transversal variable. There is no species or
sub-region dimension here; that is the essential difference between the two datasets.

```{r}
#| label: data-variables

fs_data |>
  distinct(variable_group, variable_code, variable_name, unit) |>
  arrange(variable_group, variable_code) |>
  gt(groupname_col = "variable_group")
```

## Are they the same data?

They are different tables that overlap in one variable. `FS data` reports the gross value of
landings for a fleet segment as `totlandginc`; the landings sheets report the same economic concept
as `totvallandg`, broken down by species and sub-region. The two agree closely without being
identical.

```{r}
#| label: spot-check

value_by_year <- bind_rows(
  fs_data |>
    filter(variable_code == "totlandginc") |>
    summarise(euro = sum(value, na.rm = TRUE), .by = year) |>
    mutate(source = "FS data"),
  fs_landings |>
    filter(variable_code == "totvallandg") |>
    summarise(euro = sum(value, na.rm = TRUE), .by = year) |>
    mutate(source = "FS landings")
) |>
  pivot_wider(names_from = source, values_from = euro) |>
  mutate(difference_pct = round(100 * (`FS landings` - `FS data`) / `FS data`, 1)) |>
  arrange(year)

value_by_year |>
  gt() |>
  fmt_number(c(`FS data`, `FS landings`), decimals = 0) |>
  fmt_number(difference_pct, decimals = 1)
```

From 2009 to 2023 the two series differ by less than three percent, so the landings sheets are
effectively a disaggregation of the same reporting. They are not an exact one, because the two
templates are submitted separately by Member States and because confidentiality suppression and
clustering apply differently at the two levels of detail. The 2008 and 2024 gaps are larger for the
reasons given below.

The practical consequence for the fish provisioning method is a clean division of labour. Live
weight in kilograms exists only in the landings sheets, so the biophysical account has to come from
there. Income and expenditure exist only in `FS data`, so the net economic result has to come from
there. That is why the routine reads both.

# What the NOTES sheet says

The workbook opens with a page of caveats. These are the ones that bear on our use of it.

**Fleet segments and their names.** A fleet segment is a group of similar vessels defined by vessel
length group, main fishing technology and geo indicator, operating predominantly in one
supra-region. The `fs_name` column encodes all of those. Where segments have been clustered, all
members of the cluster share one `fs_name` and the name ends with an asterisk.

**Confidentiality.** Data for segments containing fewer than ten vessels may have been excluded
altogether or merged into a cluster. This is the origin of the gaps that the processing routine
fills with average prices.

**Currency.** All monetary values in this workbook are in nominal euro. The Annual Economic Report
published from the same data uses 2023 constant prices instead. Any comparison we make across years
needs its own deflation step, and we should state the base year explicitly.

**Provisional data.** Values for 2024, where available, may be provisional, incomplete or subject to
revision.

**Relationship to the published report.** The workbook may not match the tables in the 2025 Annual
Economic Report, which applied further checks and exclusions. Data coverage and quality issues are
not flagged in the workbook itself.

Methodology and variable definitions are documented at
[stecf.jrc.ec.europa.eu/reports](https://stecf.jrc.ec.europa.eu/reports) and
[datacollection.jrc.ec.europa.eu/data-calls](https://datacollection.jrc.ec.europa.eu/data-calls).

# Coverage, and why 2024 is not usable yet

The landings sheets look complete for every year including 2024, in the sense that almost every row
carries a value. That is misleading on its own, because 2024 has only a fifth as many rows as 2023.

```{r}
#| label: coverage-landings

fs_landings |>
  summarise(
    rows = n(),
    countries = n_distinct(country_code),
    kg = sum(value[unit == "kg"], na.rm = TRUE),
    .by = year
  ) |>
  arrange(year) |>
  gt() |>
  fmt_number(c(rows, kg), decimals = 0)
```

The economic sheet is where the problem is decisive. The resource rent approach subtracts total
expenditure from total income for each fleet segment. In 2024 no Member State has yet reported any
expenditure variable except a handful of energy cost figures.

```{r}
#| label: coverage-economic

fs_data |>
  filter(variable_group %in% c("Income", "Expenditure"), year >= 2022) |>
  summarise(reported = sum(!is.na(value)), .by = c(year, variable_group, variable_code)) |>
  pivot_wider(names_from = year, values_from = reported, names_prefix = "y") |>
  arrange(variable_group, variable_code) |>
  gt(groupname_col = "variable_group")
```

Only eleven of the twenty-two Member States reported any income for 2024, and the absent group
includes Spain, France, Italy, Portugal, Ireland and Greece.

```{r}
#| label: coverage-countries

fs_data |>
  filter(variable_group == "Income", !is.na(value), year %in% c(2023, 2024)) |>
  summarise(euro = sum(value), .by = c(country_code, year)) |>
  pivot_wider(names_from = year, values_from = euro, names_prefix = "y") |>
  mutate(pct_of_2023 = round(100 * y2024 / y2023)) |>
  arrange(pct_of_2023) |>
  gt() |>
  fmt_number(c(y2023, y2024), decimals = 0) |>
  sub_missing(missing_text = "not reported")
```

**The latest year that can be valued in full is 2023.** The 2024 landings can be reported in
physical terms with a clear provisional caveat, but they cannot carry a monetary valuation until the
next data call fills in the expenditure side.

# Regenerating the files

Download the workbook from the STECF data dissemination page, place it in `data/stecf/` under its
original name, and run the script from the project root. Nothing else in the repository needs to
change; the processing routine reads the RDS files, not the workbook.
