---
title: "Marine Ecosystem Services Processing"
subtitle: "Fish provisioning: annotated R processing routine"
format: html
toc: true
toc-depth: 3
bibliography: processing-references.bib
csl: ../../apa-6th-edition.csl
execute:
  warning: false
  message: false
---

# Introduction

This page documents the R routine that turns the Scientific, Technical and Economic Committee for
Fisheries (STECF) extraction [@stecf2024] into the fish provisioning accounts. Fish provisioning is
the ecosystem contribution to the growth of aquatic biomass captured in uncultivated production
contexts [@hainesyoung2018], and it is valued here with the resource rent approach of the SEEA
Central Framework [@un2014]. The companion page on
[methods](../marine-methods/index.qmd) states the equations; this page restates the earlier working
script in a compact, reproducible form so that the team can inspect each step, reproduce it from the
repository alone, and reuse the parts that generalise.

The routine reads two committed RDS files produced from the STECF fleet-segment workbook:
`FS_data_2008-2024.RDS` for income and expenditure, and `FS_landings_FAO_2008-2024.RDS` for
landings in kilograms and euro. Kilograms exist only in the landings file; income and expenditure
exist only in `FS data`. The conversion is documented on the
[STECF Fleet Segment Data](../../FS-landings-data/) page.

The routine has three components, and they run in this order because the second depends on the
first:

1. the **economic component**, which reduces the STECF economic variables to a single net economic
   result per fleet segment and year (Eq. 3 of the methods report);
2. the **biophysical component**, which cleans the landings table, harmonises fishing sub-regions,
   assigns each sub-region to a European sea region, and attaches the species typology;
3. the **resource rent allocation**, which distributes each fleet's net economic result across the
   sub-regions and species it landed, then fills the gaps left by fleets without usable economic
   data.

::: {.callout-important}
## 2024 is excluded from the valuation

The STECF files run to 2024, but no Member State has reported a usable expenditure series for that
year. The latest fully valuable year is 2023. The two year filters below therefore keep
`year <= reference_year` and drop 2024. Landings for 2024 can be shown in physical terms with a
provisional caveat; they cannot carry a resource rent.
:::

::: {.callout-note}
## Monetary values are 2010 euro

Allocation is done in nominal euro. A single step at the end converts resource rent to constant
2010 prices with Eurostat HICP for EU-27 [@eurostat2026hicp]. The base year is `base_year` in the
parameters chunk (set to 2010 after consultation with JRC); changing it does not touch the
allocation. Tables report both nominal and 2010 euro. Physical landings are not deflated.
:::

::: {.callout-note}
## Paths are relative to the project root

`_quarto.yml` sets `execute-dir: project`, so every path below resolves from the repository root
regardless of where the `.qmd` file sits. There is no need for `here::here()` or `setwd()`.
:::

## Preliminaries

We work in the tidyverse throughout. `readxl` reads the lookup workbooks, `sf` joins the results to
the marine region geometry, `writexl` writes the flat panel workbook, `openxlsx2` writes the
formatted results workbook in the Outputs section, and `gt` formats the inspection tables.
`writexl` cannot set number formats, column widths or borders, so it is kept only for the
unformatted panel.

```{r}
#| label: libraries

library(tidyverse)
library(readxl)
library(sf)
library(writexl)
library(openxlsx2)
library(gt)

sf_use_s2(FALSE)
```

Every input and output path is declared once, at the top, so that a reader can see the full data
footprint of the routine without scrolling. `reference_year` is the latest fully valuable year. The
two year filters use `year <= reference_year` rather than `year == reference_year`; `year` is
already part of every grouping key and every gap-filling join, so the result is a 2008 to 2023
panel with no cross-year contamination.

```{r}
#| label: parameters

reference_year <- 2023
base_year <- 2010

# Inputs
stecf_economic_file <- "data/stecf/FS_data_2008-2024.RDS"
stecf_landings_file <- "data/stecf/FS_landings_FAO_2008-2024.RDS"
country_codes_file <- "data/stecf/country_codes.csv"
species_lookup_file <- "data/lookup/3.classification_vs_OFFICIAL.xlsx"
gsa_lookup_file <- "data/lookup/Link_GSA_Marine regions_Mediterranean Sea.xlsx"
overuse_lookup_file <- "data/lookup/Overuse_Total.xlsx"
hicp_file <- "data/lookup/hicp_eu27.csv"
gis_file <- "data/gisdata/marine_regions.gpkg"

# Outputs
results_dir <- "data/outputs/results"
gis_dir <- "data/outputs/gis"
dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(gis_dir, showWarnings = FALSE, recursive = TRUE)

country_names <- readr::read_csv(country_codes_file, show_col_types = FALSE) |>
  transmute(country_code = `alpha-3`, country = name) |>
  mutate(
    country = recode(country, `Netherlands, Kingdom of the` = "Netherlands")
  )
```

# Reference tables

The classification decisions of the assessment were previously buried inside long `case_when()`
blocks or made by hand in Excel. They are collected here instead, as tables wherever a table will
do. The team can then review and amend them without reading the pipeline, and they are easy to
export for the annex of the report.

## Sub-region harmonisation

STECF reports fishing areas at a finer resolution than the accounts require, and it mixes two
naming systems: FAO fishing areas such as `37.1.1` and General Fisheries Commission for the
Mediterranean sub-areas such as `sa21`. The table below collapses the finer codes onto the code
actually used in the accounts. It reproduces the recodes of the original script, plus three
collapses Domenico asked for: `sa11.1` and `sa11.2` onto `sa11` (Mediterranean), the Baltic
subdivisions `27.3.d.28.1` and `27.3.d.28.2` onto `27.3.d.28`, and the parent ICES area `27.8`
onto `27.8.a`.

```{r}
#| label: sub-region-recodes
#| eval: true

sub_reg_recodes <- tribble(
  ~from,       ~to,
  "37.1.1",    "sa5",
  "sa21.1",    "sa21",
  "sa21.2",    "sa21",
  "sa21.3",    "sa21",
  "37.2.2",    "sa21",
  "37.1.3",    "sa21",
  "37.3.1",    "sa22",
  "27.2.a.1",  "27.2.a",
  "27.2.a.2",  "27.2.a",
  "27.2.b.1",  "27.2.b",
  "27.2.b.2",  "27.2.b",
  "27.3.a.20", "27.3.a",
  "27.3.a.21", "27.3.a",
  "27.3.c.22", "27.3.c",
  "27.3.b.23", "27.3.b",
  "27.5.a.1",  "27.5.a",
  "27.5.a.2",  "27.5.a",
  "27.5.b.1",    "27.5.b",
  "27.5.b.2",    "27.5.b",
  "27.14.b.1",   "27.14.b",
  "27.14.b.2",   "27.14.b",
  "sa11.1",      "sa11",
  "sa11.2",      "sa11",
  "27.3.d.28.1", "27.3.d.28",
  "27.3.d.28.2", "27.3.d.28",
  "27.8",        "27.8.a"
)
```

The correspondence between GFCM sub-areas and FAO Mediterranean divisions is held separately in
`data/lookup`. The pipeline does not consume it; it is the reference that justifies the
Mediterranean rows of the recode table, and it is the authority against which the sea region rules
below can be audited.

```{r}
#| label: gsa-lookup
#| eval: true

gsa_to_fao <- read_excel(gsa_lookup_file, sheet = "Regions division") |>
  rename(sub_reg = `Row Labels`, fao_division = Sea_Region)
```


## Sea region assignment

Each harmonised sub-region belongs to one of the four pan-European marine regions of the Marine
Strategy Framework Directive, or to the rest of the world when a European fleet lands a catch taken
elsewhere. The rules are a direct transcription of the original `case_when()`; wrapping them in a
function means the same classification can be reused by the overuse and blue carbon routines
without copying the list of codes again.

```{r}
#| label: fn-assign-sea-region
#| eval: true

neao_sub_regions <- c(
  "27.1.a",  "27.1.b",  "27.10.a", "27.10.b", "27.2.a",  "27.2.b",
  "27.3.a",  "27.3.c",  "27.3.b",  "27.4.a",  "27.4.b",  "27.4.c",
  "27.5.a",  "27.5.b",  "27.6.a",  "27.6.b",  "27.7.a",  "27.7.b",
  "27.7.c",  "27.7.d",  "27.7.e",  "27.7.f",  "27.7.g",  "27.7.h",
  "27.7.j",  "27.7.k",  "27.8.a",  "27.8.b",  "27.8.c",  "27.8.d",
  "27.8.e",  "27.9.a",  "27.9.b",  "27.12.a", "27.12.b", "27.12.c",
  "27.14.a", "27.14.b"
)

assign_sea_region <- function(sub_reg_clean) {
  case_when(
    sub_reg_clean == "0" ~ "unknown",
    sub_reg_clean %in% c("sa28", "sa29", "sa30") ~ "Black Sea",
    sub_reg_clean %in% paste0("sa", 1:27) ~ "Mediterranean Sea",
    str_detect(sub_reg_clean, fixed("27.3.d")) ~ "Baltic Sea",
    str_detect(sub_reg_clean, fixed("27.8")) |
      sub_reg_clean %in% neao_sub_regions ~ "NEAO",
    .default = "ROW"
  )
}
```

## Species typology

STECF identifies species by their FAO ASFIS code but says nothing about what kind of animal it is
or where in the water column it lives. The methods report fills that gap from FishBase and
SeaLifeBase [@froese2024; @palomares2024], and the result is the classification workbook already in
`data/lookup`. We derive two attributes from it: `species_group`, which separates fish from
shellfish and from species whose code could not be matched with confidence, and `fish_type`, the
aggregated water column typology of Table 1 of the methods report. Those names match the nested
columns of methods Tables 4 to 13. The workbook column `Type` holds the water-column codes
(benthic, benthopelagic, demersal, pelagic); the fish / shellfish / other split comes from the
binary flags. There is no `Type_1` column in the lookup.

```{r}
#| label: species-typology
#| eval: true

species_types_raw <- read_excel(species_lookup_file, sheet = "Classification")

species_types <- species_types_raw |>
  transmute(
    species_code,
    species_name_lookup = species_name,
    fish_type = Type,
    species_group = case_when(
      fish == 1 ~ "fish",
      shellfish == 1 ~ "shellfish",
      .default = "other"
    )
  ) |>
  distinct(species_code, .keep_all = TRUE)
```

::: {.callout-note}
## Duplicate species codes are identical pairs

The original script joined landings to this classification by hand in Excel because the join in R
produced double counting. Eight species codes appear twice in the `Classification` sheet (CAS, RKQ,
SCY, SHR, SJA, SRK, SSB, YRS). Each pair is identical on name, `Type`, and the fish / shellfish
flags, so `distinct(species_code, .keep_all = TRUE)` is a safe delete. Domenico confirmed that
dropping the duplicates is fine.
:::

```{r}
#| label: check-duplicate-species
#| eval: true

species_types_raw |>
  filter(n() > 1, .by = species_code) |>
  arrange(species_code) |>
  select(species_code, species_name, Type, fish, shellfish, Other) |>
  gt()
```

# Helper functions

These operations recur often enough to be worth naming. Everything else stays as an explicit
pipeline, because a named function that is called once hides more than it saves.

Both STECF sheets carry the same block of administrative columns that the assessment never uses.
`any_of()` rather than `all_of()` means the function works on both sheets even though they do not
carry exactly the same set.

```{r}
#| label: fn-drop-admin-columns

drop_stecf_admin_columns <- function(data) {
  data |>
    select(-any_of(c(
      "upload_date", "geo_indicator", "cluster_name", "fromtable",
      "framework", "template_name", "gear", "fishery", "activity"
    )))
}
```

Harmonising sub-regions is two steps: strip the stray whitespace that STECF sometimes carries, then
apply the recode table. We strip first so that a code entered as `27.2.a.1 ` still matches. The raw
`sub_reg` is kept alongside `sub_reg_clean` for traceability.

```{r}
#| label: fn-harmonise-sub-reg
#| eval: true

harmonise_sub_reg <- function(data, recodes = sub_reg_recodes) {
  data |>
    mutate(sub_reg_clean = str_remove_all(sub_reg, "\\s")) |>
    left_join(
      recodes,
      by = join_by(sub_reg_clean == from),
      relationship = "many-to-one"
    ) |>
    mutate(sub_reg_clean = coalesce(to, sub_reg_clean)) |>
    select(-to)
}
```

The gap filling described in Box 2 of the methods report is the same operation applied three times
at successively coarser levels of aggregation. Expressing it once as a function that takes the
grouping columns as an argument removes about sixty lines of copied code and guarantees that the
three passes stay consistent with each other.

```{r}
#| label: fn-fill-price-gaps

# Mean observed price of a species over an arbitrary set of grouping columns.
average_price <- function(reference, by) {
  reference |>
    summarise(price = mean(price, na.rm = TRUE), .by = all_of(by))
}

# Assign that mean price to records that have landings but no usable rent.
# Records that still find no reference price keep a price of zero and are
# passed on to the next, coarser pass.
fill_price_gaps <- function(gaps, reference, by, basis) {
  gaps |>
    select(-any_of(c("price", "sum_revenue", "price_basis"))) |>
    left_join(
      average_price(reference, by),
      by = by,
      relationship = "many-to-one"
    ) |>
    mutate(
      price = replace_na(price, 0),
      price_basis = if_else(price > 0, basis, NA_character_),
      sum_revenue = sum_landed * price
    )
}
```

# Economic component

## Loading the STECF economic table

The economic file reports one row per fleet segment, year and economic variable. Inactive fleets
are removed first because they land nothing and would otherwise distort the fleet averages. The
year filter keeps the panel through `reference_year` and drops 2024, which cannot be valued.

```{r}
#| label: economic-load

stecf_economic <- readRDS(stecf_economic_file) |>
  filter(fishing_tech != "INACTIVE", year <= reference_year) |>
  drop_stecf_admin_columns()
```

## Net economic result by fleet

The resource rent approach values the service as the economic surplus of the fleet, that is total
income less total costs (Eq. 3 of the methods report). STECF splits its economic variables into
groups; we keep only `Expenditure` and `Income`, sign the expenditure negatively so that both can
be summed in a single column, and aggregate to one net result per fleet segment and year.

Missing values become zero and zero rows are then dropped. That looks circular, and it is
deliberate: STECF uses a blank to mean "not reported" and the assessment treats a non-reported item
as absent rather than as a true zero, so both end up excluded from the sum.

```{r}
#| label: fn-net-economic-by-fleet

net_economic_by_fleet <- function(data) {
  data |>
    filter(variable_group %in% c("Expenditure", "Income")) |>
    mutate(value = replace_na(value, 0)) |>
    filter(value != 0) |>
    mutate(money = if_else(variable_group == "Expenditure", -value, value)) |>
    summarise(
      total_expenditure = sum(money[variable_group == "Expenditure"]),
      total_income = sum(money[variable_group == "Income"]),
      total_net = total_income + total_expenditure,
      .by = c(year, fs_name)
    )
}

fleet_net <- net_economic_by_fleet(stecf_economic)
```

An ecosystem service cannot have a negative flow, because that would describe a disservice. Box 2
of the methods report therefore splits the fleets in two: those with a positive net result carry the
valuation, and those with a negative or zero result are treated as missing and recovered later from
average prices.

```{r}
#| label: fleet-net-positive

fleet_net_positive <- fleet_net |>
  filter(total_net > 0)
```

# Biophysical component

## Loading and cleaning landings

The landings file reports quantities in several units. Only kilograms are relevant here, and rows
with a zero quantity carry no information. We rename `value` to `landed_kg` immediately, because
three different quantities are called `value` at various points of the original script and the name
stops carrying meaning. The same `year <= reference_year` bound used on the economic file keeps the
two sides of the join on the same years.

```{r}
#| label: landings-load

landings_raw <- readRDS(stecf_landings_file) |>
  filter(year <= reference_year) |>
  drop_stecf_admin_columns() |>
  filter(unit == "kg", value != 0) |>
  rename(landed_kg = value)
```

## Harmonising regions and attaching the typology

Sub-region harmonisation, sea region assignment and the species join all happen in one pipeline.
The species join is the step that was previously done by hand in Excel; `relationship =
"many-to-one"` makes `dplyr` fail loudly if the lookup ever regains duplicate codes, which is the
guard the manual workflow lacked.

```{r}
#| label: landings-clean

landings <- landings_raw |>
  harmonise_sub_reg() |>
  mutate(Sea_reg = assign_sea_region(sub_reg_clean)) |>
  left_join(
    species_types,
    by = join_by(species_code),
    relationship = "many-to-one"
  )
```

Three checks are worth running every time the extraction is refreshed. The first lists species codes
present in the landings but absent from the classification workbook; those rows would otherwise
reach the accounting tables with a missing fish type. The second compares the species name STECF
reports against the name the classification workbook holds for the same code, which is how a
mis-assigned code shows itself. The third lists sub-regions that fell through to `ROW`, the
catch-all category and therefore the place where a new or misspelled STECF code will hide.

```{r}
#| label: check-unmatched

landings |>
  filter(year == reference_year, is.na(fish_type)) |>
  distinct(species_code, species_name) |>
  arrange(species_code) |>
  gt()

landings |>
  filter(year == reference_year, species_name != species_name_lookup) |>
  distinct(species_code, species_name, species_name_lookup) |>
  arrange(species_code) |>
  gt()

landings |>
  filter(year == reference_year, Sea_reg == "ROW") |>
  summarise(landed_kg = sum(landed_kg), .by = c(sub_reg, sub_reg_clean)) |>
  arrange(desc(landed_kg)) |>
  gt()
```

A fourth check needs no STECF data at all, so it is the one to look at first. Every sub-area listed
in the GFCM correspondence workbook is Mediterranean by definition; any of them that the recode and
classification rules fail to place in the Mediterranean is a gap in those rules.

```{r}
#| label: check-mediterranean-coverage
#| eval: true

gsa_to_fao |>
  harmonise_sub_reg() |>
  mutate(assigned = assign_sea_region(sub_reg_clean)) |>
  filter(assigned != "Mediterranean Sea") |>
  gt()
```

::: {.callout-note}
## `sa11.1` and `sa11.2` are recoded to `sa11`

`assign_sea_region()` matches Mediterranean sub-areas against the literal list `sa1` to `sa27`. The
decimal codes `sa11.1` and `sa11.2` would otherwise fall through to `ROW`. They are GFCM
Mediterranean areas, so the recode table now maps both onto `sa11`. That check should return no
rows. The move is a few tens of tonnes in 2023; Mediterranean and ROW totals therefore differ
slightly from the shapefile `Tonnes_23` attributes, which were built before this recode.
:::

The 2023 landings, once harmonised and assigned to a sea region, must still match the shapefile
`Tonnes_23` attributes for NEAO, the Baltic and the Black Sea. Mediterranean and ROW shift by the
tonnage of `sa11.1` and `sa11.2`, which now sit in the Mediterranean.

```{r}
#| label: check-2023-tonnes-anchor

anchor_2023 <- tribble(
  ~Sea_reg,             ~anchor_tonnes,
  "NEAO",                     1925814,
  "Baltic Sea",                436672,
  "Mediterranean Sea",         308722,
  "Black Sea",                   9961,
  "ROW",                       713131
)

sa11_tonnes <- landings_raw |>
  filter(
    year == reference_year,
    str_remove_all(sub_reg, "\\s") %in% c("sa11.1", "sa11.2")
  ) |>
  summarise(sa11_tonnes = sum(landed_kg) / 1000) |>
  pull(sa11_tonnes)

anchor_expected <- anchor_2023 |>
  mutate(
    expected_tonnes = case_when(
      Sea_reg == "Mediterranean Sea" ~ anchor_tonnes + sa11_tonnes,
      Sea_reg == "ROW" ~ anchor_tonnes - sa11_tonnes,
      .default = anchor_tonnes
    )
  )

anchor_check <- landings |>
  filter(year == reference_year) |>
  summarise(landed_tonnes = sum(landed_kg) / 1000, .by = Sea_reg) |>
  inner_join(anchor_expected, by = "Sea_reg") |>
  mutate(
    difference_tonnes = landed_tonnes - expected_tonnes,
    match = abs(difference_tonnes) < 0.6
  ) |>
  arrange(Sea_reg)

anchor_check |>
  select(Sea_reg, landed_tonnes, expected_tonnes, difference_tonnes, match) |>
  gt() |>
  fmt_number(c(landed_tonnes, expected_tonnes, difference_tonnes), decimals = 1) |>
  cols_label(
    landed_tonnes = "Landings (tonnes)",
    expected_tonnes = "Expected (tonnes)",
    difference_tonnes = "Difference"
  ) |>
  tab_source_note(
    paste0(
      "Published shapefile Tonnes_23 is the starting point. sa11.1 + sa11.2 = ",
      round(sa11_tonnes, 1),
      " t moved from ROW to the Mediterranean."
    )
  )

stopifnot(
  "2023 sea-region tonnes must match the published shapefile anchors after the sa11 recode" =
    nrow(anchor_check) == nrow(anchor_expected) && all(anchor_check$match)
)
```

# Resource rent allocation

## Matching landings to fleet economics

Landings and fleet economics meet on the fleet segment name and the year. The join is
many-to-one because a fleet has one net result but many landing records. Records that find no
positive net result are the gap set of Box 2; they keep their quantities and are valued later from
average prices.

```{r}
#| label: landings-matched

landings_matched <- landings |>
  left_join(
    fleet_net_positive,
    by = join_by(fs_name, year),
    relationship = "many-to-one"
  )

landings_valued_input <- landings_matched |>
  filter(!is.na(total_net))

landings_without_rent <- landings_matched |>
  filter(is.na(total_net))
```

## Distributing the net result

A fleet earns one net result but fishes several sub-regions and several species, and STECF cannot
attribute cost or revenue any more finely than the fleet. The assessment therefore distributes the
net result in proportion to the quantities landed, in two nested steps: first across the
sub-regions the fleet worked (Eq. 4 and Eq. 5), then across the species it landed in each of those
sub-regions (Eq. 6 and Eq. 7).

The original script carried these weights as percentages and divided by one hundred again at each
use. We keep them as shares, which is arithmetically identical and removes four opportunities for a
misplaced factor of one hundred.

```{r}
#| label: rent-allocation

landings_valued <- landings_valued_input |>
  mutate(
    fleet_landed_kg = sum(landed_kg, na.rm = TRUE),
    .by = c(year, fs_name)
  ) |>
  mutate(
    fleet_area_landed_kg = sum(landed_kg, na.rm = TRUE),
    .by = c(year, fs_name, sub_reg_clean)
  ) |>
  mutate(
    # Eq. 4: share of the fleet's catch taken in this sub-region
    share_fleet_area = fleet_area_landed_kg / fleet_landed_kg,
    # Eq. 5: net result attributable to the fleet in this sub-region
    net_fleet_area = total_net * share_fleet_area,
    # Eq. 6: share of that sub-regional catch made up by this species
    share_species_area = landed_kg / fleet_area_landed_kg,
    # Eq. 7: net result attributable to the species in this sub-region
    net_species = net_fleet_area * share_species_area
  )
```

## Observed prices

The accounts report one figure per species, landing country and sub-region, so the fleet dimension
is aggregated away here. Dividing the summed net result by the summed quantity gives the implied
resource rent per kilogram, which is the price used to value the gap records in the next section.

```{r}
#| label: rent-prices

accounting_keys <- c(
  "year", "country_code", "sub_reg_clean", "Sea_reg",
  "species_code", "species_name", "species_group", "fish_type"
)

rent_prices <- landings_valued |>
  summarise(
    sum_landed = sum(landed_kg, na.rm = TRUE),
    sum_revenue = sum(net_species, na.rm = TRUE),
    .by = all_of(accounting_keys)
  ) |>
  mutate(
    price = if_else(sum_landed > 0, sum_revenue / sum_landed, NA_real_),
    price_basis = "observed rent"
  )
```

::: {.callout-note}
## Two deliberate differences from the original script

The original produced this table with `mutate()` followed by `distinct()`, which left one row per
underlying fleet record rather than one row per accounting key; because the raw `sub_reg` was among
the columns kept, sub-regions that had been collapsed by the recode table survived as separate rows
carrying the same group total, and the final aggregation added them together. Using `summarise()`
over an explicit key list produces exactly one row per group and removes that double counting.

The original also guarded the division against a zero in the individual record's own net value
rather than in the group's total quantity, which set a price to `NA` for records that had a valid
group price. The guard here is on `sum_landed`, which is the quantity actually in the denominator.
:::

## Gap set

Landings from fleets without a positive net result are aggregated onto the same accounting keys so
that the two halves of the dataset share one grain. They arrive with quantities and no revenue.

```{r}
#| label: price-gaps

price_gaps <- landings_without_rent |>
  summarise(
    sum_landed = sum(landed_kg, na.rm = TRUE),
    .by = all_of(accounting_keys)
  )
```

## Filling the gaps

Box 2 of the methods report fills the gaps in three passes, each one coarser than the last: the
average price of the species in the same country, then in the same sub-region, then across Europe.
Each pass keeps what it could value and hands the remainder to the next. Anything still unvalued
after the third pass has no observed price anywhere in the dataset for that species.

```{r}
#| label: gap-filling

# Pass 1: average price of the species in the same landing country
filled_country <- fill_price_gaps(
  price_gaps,
  rent_prices,
  by = c("year", "species_code", "species_name", "country_code"),
  basis = "country average"
)

# Pass 2: average price of the species in the same sub-region
filled_sub_reg <- filled_country |>
  filter(price == 0) |>
  fill_price_gaps(
    rent_prices,
    by = c("year", "species_code", "species_name", "sub_reg_clean"),
    basis = "sub-region average"
  )

# Pass 3: average price of the species anywhere in the dataset
filled_europe <- filled_sub_reg |>
  filter(price == 0) |>
  fill_price_gaps(
    rent_prices,
    by = c("year", "species_code", "species_name"),
    basis = "European average"
  )
```

The passes overlap by construction, so the first two contribute only the records they actually
valued. The third contributes everything that reaches it, including the records it could not value;
those carry a quantity and a revenue of zero, which is what keeps the physical account complete
even where the economic data does not reach. They are also extracted on their own, because their
size is a useful measure of that reach.

```{r}
#| label: gap-filling-bind

accounting_records <- bind_rows(
  rent_prices,
  filled_country |> filter(price > 0),
  filled_sub_reg |> filter(price > 0),
  filled_europe |> mutate(price_basis = replace_na(price_basis, "unvalued"))
)

unvalued_records <- filled_europe |>
  filter(price == 0)
```

# Final dataset

The four sources are aggregated once more onto the accounting keys, because a species, country and
sub-region combination can receive quantities from both a positive-rent fleet and a gap-filled
fleet. The final price is the revenue-weighted average that results, and the report's units are
tonnes and thousands of euro.

`price_basis` is collapsed into a readable label rather than dropped, so each row of the published
table carries the provenance of its own valuation. Note that both unit conversions are applied to
the aggregated columns; the original script converted quantities from the aggregate but converted
revenue from the pre-aggregation column, so the two published units were computed on different
bases.

```{r}
#| label: fish-provisioning

fish_provisioning <- accounting_records |>
  summarise(
    landed_kg = sum(sum_landed, na.rm = TRUE),
    revenue_eur = sum(sum_revenue, na.rm = TRUE),
    price_basis = str_c(sort(unique(price_basis)), collapse = " + "),
    .by = all_of(accounting_keys)
  ) |>
  mutate(
    price_eur_per_kg = if_else(
      landed_kg > 0, revenue_eur / landed_kg, NA_real_
    ),
    landed_tonnes = landed_kg / 1000,
    revenue_thousand_eur = revenue_eur / 1000
  ) |>
  arrange(year, Sea_reg, country_code, species_code)
```

Deflation is applied here, after the accounts exist, and nowhere else. Eurostat HICP for EU-27,
all-items, annual average index, is read from a committed lookup so the site builds without a
network call [@eurostat2026hicp]. The factor is `HICP_t / HICP_base`. Resource rent in constant
euro is nominal rent divided by that factor. The base year is 2010.

```{r}
#| label: deflate

hicp <- read_csv(hicp_file, show_col_types = FALSE)
hicp_base <- hicp$hicp_index[hicp$year == base_year]
if (length(hicp_base) != 1 || is.na(hicp_base) || hicp_base <= 0) {
  stop("HICP lookup has no usable index for base_year = ", base_year)
}

fish_provisioning <- fish_provisioning |>
  left_join(hicp, by = "year") |>
  mutate(
    deflator = hicp_index / hicp_base,
    revenue_eur_constant = revenue_eur / deflator,
    revenue_thousand_eur_constant = revenue_eur_constant / 1000,
    price_eur_per_kg_constant = if_else(
      landed_kg > 0, revenue_eur_constant / landed_kg, NA_real_
    )
  )

hicp |>
  mutate(
    deflator = hicp_index / hicp_base,
    base = if_else(year == base_year, "base", "")
  ) |>
  gt() |>
  cols_label(
    year = "Year",
    hicp_index = "HICP (2015 = 100)",
    deflator = paste0("Deflator (", base_year, " = 1)"),
    base = ""
  ) |>
  fmt_number(hicp_index, decimals = 2) |>
  fmt_number(deflator, decimals = 4) |>
  tab_source_note(
    "Eurostat prc_hicp_aind, EU-27 (from 2020), CP00, annual average index."
  )
```

Overuse is applied here, after allocation and deflation, and is not folded into Eq. 3 to 7. The
official SUT records the actual flow. The FSI overlay uses the integer scores in
`data/lookup/Overuse_Total.xlsx`, which is the 2018 worked example: scores differ by sea and cover
only the assessed species of methods Table 19. Table 18 converts that integer to an overuse share
of actual flow. Unassessed species are left blank, not treated as sustainable. Aquaculture ASI is
out of scope.

```{r}
#| label: overuse

overuse_share_from_score <- function(score) {
  case_when(
    is.na(score) ~ NA_real_,
    score < 0 ~ 1,
    score == 0 ~ NA_real_,
    score == 1 ~ 0.9,
    score == 2 ~ 0.7,
    score == 3 ~ 0.5,
    score == 4 ~ 0.4,
    score == 5 ~ 0.3,
    score == 6 ~ 0.2,
    score == 7 ~ 0.1,
    score >= 8 ~ 0
  )
}

read_overuse_sheet <- function(sheet, sea) {
  read_excel(overuse_lookup_file, sheet = sheet) |>
    rename(species_code = 1) |>
    transmute(
      Sea_reg = sea,
      species_code,
      fsi_common_name = `Common name`,
      fsi_initial_score = `Initial Score`,
      fsi_score = Score,
      fsi_risk_lookup = `Risk level`
    )
}

overuse_scores <- bind_rows(
  read_overuse_sheet("Baltic Sea", "Baltic Sea"),
  read_overuse_sheet("Black Sea", "Black Sea"),
  read_overuse_sheet("Med Sea", "Mediterranean Sea"),
  read_overuse_sheet("NEAO", "NEAO")
) |>
  mutate(overuse_share = overuse_share_from_score(fsi_score))

fish_provisioning <- fish_provisioning |>
  left_join(overuse_scores, by = c("Sea_reg", "species_code")) |>
  mutate(
    overuse_tonnes = if_else(
      !is.na(overuse_share), landed_tonnes * overuse_share, NA_real_
    ),
    sustainable_tonnes = if_else(
      !is.na(overuse_share), landed_tonnes * (1 - overuse_share), NA_real_
    ),
    overuse_thousand_eur = if_else(
      !is.na(overuse_share), revenue_thousand_eur * overuse_share, NA_real_
    ),
    overuse_thousand_eur_constant = if_else(
      !is.na(overuse_share),
      revenue_thousand_eur_constant * overuse_share,
      NA_real_
    )
  )

overuse_by_sea_year <- fish_provisioning |>
  filter(!is.na(overuse_share)) |>
  summarise(
    assessed_tonnes = sum(landed_tonnes),
    overuse_tonnes = sum(overuse_tonnes),
    assessed_thousand_eur = sum(revenue_thousand_eur),
    overuse_thousand_eur = sum(overuse_thousand_eur),
    assessed_thousand_eur_constant = sum(revenue_thousand_eur_constant),
    overuse_thousand_eur_constant = sum(overuse_thousand_eur_constant),
    .by = c(year, Sea_reg)
  ) |>
  mutate(
    share_tonnes = overuse_tonnes / assessed_tonnes,
    share_eur = overuse_thousand_eur / assessed_thousand_eur
  )

overuse_by_sea_year |>
  filter(year == 2018) |>
  arrange(Sea_reg) |>
  gt() |>
  fmt_number(
    c(
      assessed_tonnes, overuse_tonnes,
      assessed_thousand_eur, overuse_thousand_eur
    ),
    decimals = 2
  ) |>
  fmt_percent(c(share_tonnes, share_eur), decimals = 0) |>
  tab_header(title = "Overuse overlay, assessed species only, 2018") |>
  tab_source_note(
    "Shares apply only to species with an FSI score in that sea. This is not a column of the official SUT. Scores are the 2018 worked example and are held constant across years until a new stock-assessment vintage arrives."
  )
```

# Accounting tables

The methods report prints nested supply and use tables, not the compact sea-by-country summaries.
Supply rows are FAO/GFCM fishing areas (`sub_reg_clean`), labelled EEZ as in the report; they are
the spatial unit of supply, not legal exclusive economic zones. Use rows are the Member State of
landing. Columns are `species_group` crossed with `fish_type`, produced with `pivot_wider()` so
the table comes out of the pipeline in the same shape as methods Tables 10 to 13. Unmatched
species sit in Others / Not identified. Empty fish-type combinations are filled with zero.
Supply totals equal use totals at every grain. The 2023 Baltic tables below are the worked
example of chapter 4; the Excel workbook repeats them for every sea and year, in tonnes, thousand
nominal euro, and thousand 2010 euro.

```{r}
#| label: sut-helpers
#| output: false

sea_order <- c(
  "Baltic Sea", "Black Sea", "Mediterranean Sea", "NEAO", "ROW", "unknown"
)

sut_value_cols <- c(
  "fish_benthopelagic", "fish_demersal", "fish_pelagic",
  "shellfish_benthic", "shellfish_benthopelagic",
  "shellfish_demersal", "shellfish_pelagic",
  "other_not_identified"
)

sut_type_labels <- c(
  fish_benthopelagic = "benthopelagic",
  fish_demersal = "demersal",
  fish_pelagic = "pelagic",
  shellfish_benthic = "benthic",
  shellfish_benthopelagic = "benthopelagic",
  shellfish_demersal = "demersal",
  shellfish_pelagic = "pelagic",
  other_not_identified = "Not identified",
  Total = "Total"
)

prepare_sut_types <- function(data) {
  data |>
    mutate(
      species_group = case_when(
        species_group == "fish" ~ "fish",
        species_group == "shellfish" ~ "shellfish",
        .default = "other"
      ),
      fish_type = case_when(
        species_group == "other" ~ "not_identified",
        is.na(fish_type) ~ "not_identified",
        .default = fish_type
      ),
      species_group = factor(species_group, c("fish", "shellfish", "other")),
      fish_type = factor(
        fish_type,
        c("benthic", "benthopelagic", "demersal", "pelagic", "not_identified")
      )
    )
}

nested_sut <- function(data, row_var, value_var) {
  wide <- data |>
    prepare_sut_types() |>
    summarise(
      value = sum(.data[[value_var]], na.rm = TRUE),
      .by = all_of(c(row_var, "species_group", "fish_type"))
    ) |>
    pivot_wider(
      names_from = c(species_group, fish_type),
      values_from = value,
      names_sep = "_",
      values_fill = 0,
      names_expand = TRUE
    )

  missing_cols <- setdiff(sut_value_cols, names(wide))
  for (col in missing_cols) {
    wide[[col]] <- 0
  }

  wide |>
    mutate(Total = rowSums(across(all_of(sut_value_cols)), na.rm = TRUE)) |>
    filter(Total != 0) |>
    select(all_of(row_var), all_of(sut_value_cols), Total) |>
    arrange(across(all_of(row_var)))
}

add_total_row <- function(tbl, row_var) {
  totals <- tbl |>
    summarise(across(where(is.numeric), \(x) sum(x, na.rm = TRUE))) |>
    mutate(!!row_var := "Total", .before = 1)
  bind_rows(tbl, totals)
}

sea_sut <- function(year, sea, side, value_var, data = fish_provisioning) {
  slice <- data |> filter(year == .env$year, Sea_reg == .env$sea)
  if (side == "supply") {
    nested_sut(slice, "sub_reg_clean", value_var) |>
      add_total_row("sub_reg_clean") |>
      rename(EEZ = sub_reg_clean)
  } else {
    slice |>
      left_join(country_names, by = "country_code") |>
      mutate(country = coalesce(country, country_code)) |>
      nested_sut("country", value_var) |>
      add_total_row("country") |>
      rename(Country = country)
  }
}

europe_supply_sut <- function(year, value_var, data = fish_provisioning) {
  data |>
    filter(year == .env$year) |>
    nested_sut("Sea_reg", value_var) |>
    mutate(Sea_reg = factor(Sea_reg, levels = sea_order)) |>
    arrange(Sea_reg) |>
    mutate(Sea_reg = as.character(Sea_reg)) |>
    add_total_row("Sea_reg") |>
    rename(`Fishing region` = Sea_reg)
}

europe_use_sut <- function(year, value_var, data = fish_provisioning) {
  data |>
    filter(year == .env$year) |>
    left_join(country_names, by = "country_code") |>
    mutate(country = coalesce(country, country_code)) |>
    nested_sut("country", value_var) |>
    add_total_row("country") |>
    rename(Country = country)
}

europe_origin_sut <- function(year, value_var, data = fish_provisioning) {
  data |>
    filter(year == .env$year) |>
    left_join(country_names, by = "country_code") |>
    mutate(country = coalesce(country, country_code)) |>
    summarise(value = sum(.data[[value_var]], na.rm = TRUE), .by = c(country, Sea_reg)) |>
    mutate(Sea_reg = factor(Sea_reg, levels = sea_order)) |>
    pivot_wider(
      names_from = Sea_reg,
      values_from = value,
      values_fill = 0,
      names_expand = TRUE
    ) |>
    mutate(Total = rowSums(across(all_of(sea_order)), na.rm = TRUE)) |>
    arrange(country) |>
    add_total_row("country") |>
    rename(Country = country)
}

sut_gt <- function(tbl, title, note) {
  row_col <- names(tbl)[1]
  tbl |>
    gt(rowname_col = row_col) |>
    cols_label(.list = as.list(sut_type_labels)) |>
    tab_spanner("Fish", c(fish_benthopelagic, fish_demersal, fish_pelagic)) |>
    tab_spanner(
      "Shellfish",
      c(
        shellfish_benthic, shellfish_benthopelagic,
        shellfish_demersal, shellfish_pelagic
      )
    ) |>
    tab_spanner("Others", other_not_identified) |>
    fmt_number(decimals = 2) |>
    tab_header(title = title) |>
    tab_source_note(note) |>
    tab_stubhead(label = row_col)
}

sut_identity <- function(supply, use) {
  abs(dplyr::last(supply$Total) - dplyr::last(use$Total)) < 0.05
}

eez_note <- paste(
  "EEZ is the FAO/GFCM fishing area used as the spatial unit of supply,",
  "not a legal exclusive economic zone. Country is the Member State of landing."
)
```

```{r}
#| label: supply-summary

baltic_supply_t <- sea_sut(reference_year, "Baltic Sea", "supply", "landed_tonnes")
baltic_use_t <- sea_sut(reference_year, "Baltic Sea", "use", "landed_tonnes")
baltic_supply_nom <- sea_sut(reference_year, "Baltic Sea", "supply", "revenue_thousand_eur")
baltic_use_nom <- sea_sut(reference_year, "Baltic Sea", "use", "revenue_thousand_eur")
baltic_supply_const <- sea_sut(
  reference_year, "Baltic Sea", "supply", "revenue_thousand_eur_constant"
)
baltic_use_const <- sea_sut(
  reference_year, "Baltic Sea", "use", "revenue_thousand_eur_constant"
)

stopifnot(
  sut_identity(baltic_supply_t, baltic_use_t),
  sut_identity(baltic_supply_nom, baltic_use_nom),
  sut_identity(baltic_supply_const, baltic_use_const)
)

sut_gt(
  baltic_supply_t,
  paste0("Table 10. Baltic Sea supply, physical (t), ", reference_year),
  eez_note
)
sut_gt(
  baltic_use_t,
  paste0("Table 12. Baltic Sea use, physical (t), ", reference_year),
  eez_note
)
sut_gt(
  baltic_supply_nom,
  paste0("Table 11. Baltic Sea supply, thousand nominal euro, ", reference_year),
  eez_note
)
sut_gt(
  baltic_use_nom,
  paste0("Table 13. Baltic Sea use, thousand nominal euro, ", reference_year),
  eez_note
)
sut_gt(
  baltic_supply_const,
  paste0(
    "Baltic Sea supply, thousand ", base_year, " euro, ", reference_year
  ),
  eez_note
)
sut_gt(
  baltic_use_const,
  paste0("Baltic Sea use, thousand ", base_year, " euro, ", reference_year),
  eez_note
)
```

# Outputs

The original script wrote twelve intermediate workbooks to disk. We write the panel twice as a
flat deliverable, once as RDS and once as Excel, so the client can open the same table in either
tool. Observed prices and unvalued catch stay as RDS. The third workbook is the report-style SUT
file: one sheet per year, tables stacked with a blank row between them. Each nested table has the
column layout of methods Tables 10 to 13. Europe-wide Tables 2 to 9 sit at the top of the sheet;
the per-sea nested SUTs follow. A Checks sheet records the supply-use identity at every grain.

```{r}
#| label: write-outputs

saveRDS(fish_provisioning, file.path(results_dir, "fish_provisioning.RDS"))
saveRDS(rent_prices, file.path(results_dir, "rent_prices.RDS"))
saveRDS(unvalued_records, file.path(results_dir, "unvalued_records.RDS"))

writexl::write_xlsx(
  list(fish_provisioning = fish_provisioning),
  path = file.path(results_dir, "fish_provisioning.xlsx")
)

result_years <- sort(unique(fish_provisioning$year), decreasing = TRUE)
nested_seas <- sea_order
col_width <- 10
number_format <- "#,##0.00"

metric_specs <- tribble(
  ~key, ~value_var, ~unit_label,
  "physical", "landed_tonnes", "physical terms (t)",
  "nominal", "revenue_thousand_eur", "monetary terms (1,000 EUR, current prices)",
  "constant", "revenue_thousand_eur_constant",
  paste0("monetary terms (1,000 EUR, constant ", base_year, " prices)")
)

europe_region_table <- function(year, value_var) {
  by_sea <- fish_provisioning |>
    filter(year == .env$year) |>
    summarise(value = sum(.data[[value_var]], na.rm = TRUE), .by = Sea_reg) |>
    mutate(Sea_reg = factor(Sea_reg, levels = sea_order)) |>
    complete(Sea_reg, fill = list(value = 0)) |>
    arrange(Sea_reg) |>
    mutate(Sea_reg = as.character(Sea_reg))
  supply <- by_sea |>
    pivot_wider(names_from = Sea_reg, values_from = value) |>
    mutate(Total = rowSums(across(everything()), na.rm = TRUE)) |>
    mutate(SUT = "Supply", .before = 1)
  use <- supply |>
    mutate(SUT = "Use") |>
    mutate(across(-c(SUT, Total), \(v) NA_real_))
  bind_rows(supply, use)
}

write_banner_block <- function(wb, sheet, start_row, n_cols, title, year, grain) {
  lines <- c(
    "European Commission, Joint Research Centre (JRC)",
    title,
    paste0(
      "Year ", year, ". ", grain,
      " FAO/GFCM fishing areas are labelled EEZ as the spatial unit of supply; ",
      "they are not legal exclusive economic zones. Constant prices use ",
      base_year, " as the base year (Eurostat HICP, EU-27)."
    )
  )
  for (i in 1:3) {
    row <- start_row + i - 1L
    wb <- wb_add_data(
      wb, sheet, x = lines[[i]],
      start_row = row, start_col = 1, col_names = FALSE
    )
    wb <- wb_merge_cells(wb, sheet, dims = wb_dims(rows = row, cols = 1:n_cols))
    wb <- wb_add_cell_style(
      wb, sheet,
      dims = wb_dims(rows = row, cols = 1),
      wrap_text = TRUE, vertical = "center"
    )
  }
  wb_set_row_heights(wb, sheet, rows = start_row:(start_row + 2L), heights = 18)
}

write_nested_headers <- function(wb, sheet, header_row, row_label) {
  group_row <- header_row
  type_row <- header_row + 1L
  group_vals <- c(
    row_label, "Fish", "Fish", "Fish",
    "Shellfish", "Shellfish", "Shellfish", "Shellfish",
    "Others", "Total"
  )
  type_vals <- c(
    row_label, "benthopelagic", "demersal", "pelagic",
    "benthic", "benthopelagic", "demersal", "pelagic",
    "Not identified", "Total"
  )
  wb <- wb_add_data(
    wb, sheet, x = matrix(group_vals, nrow = 1),
    start_row = group_row, start_col = 1, col_names = FALSE
  )
  wb <- wb_add_data(
    wb, sheet, x = matrix(type_vals, nrow = 1),
    start_row = type_row, start_col = 1, col_names = FALSE
  )
  wb <- wb_merge_cells(wb, sheet, dims = wb_dims(rows = group_row:type_row, cols = 1))
  wb <- wb_merge_cells(wb, sheet, dims = wb_dims(rows = group_row, cols = 2:4))
  wb <- wb_merge_cells(wb, sheet, dims = wb_dims(rows = group_row, cols = 5:8))
  header_dims <- wb_dims(rows = group_row:type_row, cols = 1:10)
  wb <- wb_add_font(wb, sheet, dims = header_dims, bold = TRUE)
  wb <- wb_add_cell_style(
    wb, sheet, dims = header_dims,
    wrap_text = TRUE, horizontal = "center", vertical = "center"
  )
  wb <- wb_add_border(
    wb, sheet,
    dims = wb_dims(rows = type_row, cols = 1:10),
    bottom_border = "thin",
    top_border = "none",
    left_border = "none",
    right_border = "none"
  )
  wb <- wb_set_row_heights(wb, sheet, rows = group_row:type_row, heights = 22)
  wb
}

write_data_block <- function(wb, sheet, data, data_start) {
  n_cols <- ncol(data)
  data_end <- data_start + nrow(data) - 1L
  wb <- wb_add_data(
    wb, sheet, x = data,
    start_row = data_start, start_col = 1,
    col_names = FALSE, na = "", apply_cell_style = FALSE
  )
  wb <- wb_add_numfmt(
    wb, sheet,
    dims = wb_dims(rows = data_start:data_end, cols = 2:n_cols),
    numfmt = number_format
  )
  total_row <- data_end
  total_dims <- wb_dims(rows = total_row, cols = 1:n_cols)
  wb <- wb_add_font(wb, sheet, dims = total_dims, bold = TRUE)
  wb <- wb_add_border(
    wb, sheet, dims = total_dims,
    top_border = "thin",
    bottom_border = "thin",
    left_border = "none",
    right_border = "none"
  )
  wb <- wb_set_row_heights(wb, sheet, rows = total_row, heights = 18)
  wb <- wb_set_col_widths(wb, sheet, cols = 1:n_cols, widths = col_width)
  wb
}

write_nested_table <- function(wb, sheet, start_row, title, year, grain, data) {
  n_cols <- ncol(data)
  wb <- write_banner_block(wb, sheet, start_row, n_cols, title, year, grain)
  header_row <- start_row + 4L
  wb <- write_nested_headers(wb, sheet, header_row, names(data)[1])
  data_start <- header_row + 2L
  wb <- write_data_block(wb, sheet, data, data_start)
  list(wb = wb, next_row = data_start + nrow(data) + 1L)
}

write_simple_table <- function(wb, sheet, start_row, title, year, grain, data) {
  n_cols <- ncol(data)
  wb <- write_banner_block(wb, sheet, start_row, n_cols, title, year, grain)
  header_row <- start_row + 4L
  headers <- names(data)
  wb <- wb_add_data(
    wb, sheet, x = matrix(headers, nrow = 1),
    start_row = header_row, start_col = 1, col_names = FALSE
  )
  header_dims <- wb_dims(rows = header_row, cols = 1:n_cols)
  wb <- wb_add_font(wb, sheet, dims = header_dims, bold = TRUE)
  wb <- wb_add_cell_style(
    wb, sheet, dims = header_dims,
    wrap_text = TRUE, horizontal = "center", vertical = "center"
  )
  wb <- wb_add_border(
    wb, sheet, dims = header_dims,
    bottom_border = "thin",
    top_border = "none",
    left_border = "none",
    right_border = "none"
  )
  wb <- wb_set_row_heights(wb, sheet, rows = header_row, heights = 36)
  wb <- write_data_block(wb, sheet, data, header_row + 1L)
  list(wb = wb, next_row = header_row + 1L + nrow(data) + 1L)
}

identity_rows <- list()
append_identity <- function(year, grain, metric, supply, use) {
  identity_rows[[length(identity_rows) + 1L]] <<- tibble(
    year = year,
    grain = grain,
    metric = metric,
    supply_total = dplyr::last(supply$Total),
    use_total = dplyr::last(use$Total),
    difference = dplyr::last(supply$Total) - dplyr::last(use$Total),
    match = abs(dplyr::last(supply$Total) - dplyr::last(use$Total)) < 0.05
  )
}

wb <- wb_workbook()
wb <- wb_add_worksheet(wb, "Readme")
readme_notes <- c(
  "European Commission, Joint Research Centre (JRC)",
  "Fish provisioning ecosystem service accounts, supply and use tables",
  paste0(
    "Base year ", base_year,
    ". Resource rent in thousand euro. Constant prices use Eurostat HICP for EU-27."
  ),
  "One sheet per year. Tables are stacked, with at least one blank row between them.",
  "Nested columns follow methods Tables 4 to 7 and 10 to 13: Fish (benthopelagic, demersal, pelagic), Shellfish (benthic, benthopelagic, demersal, pelagic), Others (Not identified), Total.",
  "Supply rows labelled EEZ are FAO/GFCM fishing areas (sub_reg_clean), the spatial unit of supply, not legal exclusive economic zones.",
  "Use rows are the STECF Member State of landing, shown as the ISO name.",
  "Europe tables at the top of each year sheet correspond to methods Tables 2 to 9. Per-sea nested tables correspond to Tables 10 to 13 and annexes 1 to 15 / 16 to 35.",
  "Supply Total equals use Total at every grain; see the Checks sheet.",
  "Overuse is a sustainability overlay on assessed species only. It is not a column of the official SUT. Scores come from the 2018 FSI worked example and are held constant across years.",
  "The machine-readable panel is fish_provisioning.xlsx / fish_provisioning.RDS.",
  "Source: STECF 2025 Fleet Economic data call."
)
wb <- wb_add_data(wb, "Readme", x = readme_notes, col_names = FALSE)
wb <- wb_add_cell_style(
  wb, "Readme",
  dims = wb_dims(rows = seq_along(readme_notes), cols = 1),
  wrap_text = TRUE
)
wb <- wb_set_col_widths(wb, "Readme", cols = 1, widths = 80)

for (y in result_years) {
  year_name <- as.character(y)
  wb <- wb_add_worksheet(wb, year_name)
  current_row <- 1L

  for (i in seq_len(nrow(metric_specs))) {
    spec <- metric_specs[i, ]
    region_tbl <- europe_region_table(y, spec$value_var)
    written <- write_simple_table(
      wb, year_name, current_row,
      paste0("SUT of fish provisioning in ", spec$unit_label),
      y,
      "Europe. Fishing region as ecosystem type; use is the fisheries / primary-sector total.",
      region_tbl
    )
    wb <- written$wb
    current_row <- written$next_row

    europe_supply <- europe_supply_sut(y, spec$value_var)
    europe_use <- europe_use_sut(y, spec$value_var)
    append_identity(y, "Europe nested", spec$key, europe_supply, europe_use)

    written <- write_nested_table(
      wb, year_name, current_row,
      paste0("Supply table of fish provisioning in ", spec$unit_label, ", per fish type"),
      y,
      "Europe. Rows are fishing regions.",
      europe_supply
    )
    wb <- written$wb
    current_row <- written$next_row

    written <- write_nested_table(
      wb, year_name, current_row,
      paste0("Use table of fish provisioning in ", spec$unit_label, ", per fish type"),
      y,
      "Europe. Rows are landing countries.",
      europe_use
    )
    wb <- written$wb
    current_row <- written$next_row

    origin_tbl <- europe_origin_sut(y, spec$value_var)
    written <- write_simple_table(
      wb, year_name, current_row,
      paste0("Use table of fish provisioning in ", spec$unit_label, ", per fishing region"),
      y,
      "Europe. Landing country by fishing region of origin.",
      origin_tbl
    )
    wb <- written$wb
    current_row <- written$next_row
  }

  for (sea in nested_seas) {
    sea_rows <- fish_provisioning$Sea_reg == sea & fish_provisioning$year == y
    if (!any(sea_rows)) {
      next
    }
    for (i in seq_len(nrow(metric_specs))) {
      spec <- metric_specs[i, ]
      supply_tbl <- sea_sut(y, sea, "supply", spec$value_var)
      use_tbl <- sea_sut(y, sea, "use", spec$value_var)
      append_identity(y, sea, spec$key, supply_tbl, use_tbl)

      written <- write_nested_table(
        wb, year_name, current_row,
        paste0(
          "Supply table of fish provisioning in ", sea, " in ",
          spec$unit_label, " per fish type"
        ),
        y,
        paste0(sea, ". Supply rows are FAO/GFCM fishing areas, labelled EEZ."),
        supply_tbl
      )
      wb <- written$wb
      current_row <- written$next_row

      written <- write_nested_table(
        wb, year_name, current_row,
        paste0(
          "Use table of fish provisioning in ", sea, " in ",
          spec$unit_label, " per fish type"
        ),
        y,
        paste0(sea, ". Use rows are landing countries."),
        use_tbl
      )
      wb <- written$wb
      current_row <- written$next_row
    }
  }

  wb <- wb_freeze_pane(
    wb, year_name,
    first_active_row = 6L, first_active_col = 2L
  )
}

identity_tbl <- bind_rows(identity_rows)
stopifnot(
  "Supply Total must equal use Total at every SUT grain" = all(identity_tbl$match)
)

wb <- wb_add_worksheet(wb, "Checks")
wb <- wb_add_data(wb, "Checks", x = identity_tbl)
wb <- wb_add_numfmt(
  wb, "Checks",
  dims = wb_dims(rows = 2:(nrow(identity_tbl) + 1L), cols = 4:6),
  numfmt = number_format
)
wb <- wb_set_col_widths(wb, "Checks", cols = 1:7, widths = 18)

overuse_species <- fish_provisioning |>
  filter(!is.na(overuse_share)) |>
  summarise(
    landed_tonnes = sum(landed_tonnes),
    overuse_tonnes = sum(overuse_tonnes),
    sustainable_tonnes = sum(sustainable_tonnes),
    revenue_thousand_eur = sum(revenue_thousand_eur),
    overuse_thousand_eur = sum(overuse_thousand_eur),
    revenue_thousand_eur_constant = sum(revenue_thousand_eur_constant),
    overuse_thousand_eur_constant = sum(overuse_thousand_eur_constant),
    fsi_score = dplyr::first(fsi_score),
    overuse_share = dplyr::first(overuse_share),
    fsi_common_name = dplyr::first(fsi_common_name),
    .by = c(year, Sea_reg, species_code)
  ) |>
  arrange(year, Sea_reg, species_code)

wb <- wb_add_worksheet(wb, "Overuse")
wb <- wb_add_data(
  wb, "Overuse",
  x = c(
    "Overuse overlay on assessed species only. Not part of the official SUT.",
    "FSI scores are the 2018 worked example in data/lookup/Overuse_Total.xlsx and are held constant across years."
  ),
  col_names = FALSE
)
wb <- wb_add_data(wb, "Overuse", x = overuse_species, start_row = 4)
wb <- wb_set_col_widths(wb, "Overuse", cols = 1:12, widths = 16)

wb <- wb_add_worksheet(wb, "Comparison")
cmp_physical <- europe_supply_sut(2018, "landed_tonnes") |>
  select(`Fishing region`, Total_2018 = Total) |>
  left_join(
    europe_supply_sut(2023, "landed_tonnes") |>
      select(`Fishing region`, Total_2023 = Total),
    by = "Fishing region"
  )
cmp_constant <- europe_supply_sut(2018, "revenue_thousand_eur_constant") |>
  select(`Fishing region`, Total_2018 = Total) |>
  left_join(
    europe_supply_sut(2023, "revenue_thousand_eur_constant") |>
      select(`Fishing region`, Total_2023 = Total),
    by = "Fishing region"
  )
wb <- wb_add_data(
  wb, "Comparison",
  x = paste0(
    "Tables 14 to 15 analogue. Physical tonnes and thousand ",
    base_year, " euro. Supply totals; use equals supply at this grain."
  ),
  col_names = FALSE
)
wb <- wb_add_data(wb, "Comparison", x = "Physical (t)", start_row = 3, col_names = FALSE)
wb <- wb_add_data(wb, "Comparison", x = cmp_physical, start_row = 4)
wb <- wb_add_data(
  wb, "Comparison",
  x = paste0("Thousand ", base_year, " euro"),
  start_row = 4 + nrow(cmp_physical) + 3L,
  col_names = FALSE
)
wb <- wb_add_data(
  wb, "Comparison",
  x = cmp_constant,
  start_row = 4 + nrow(cmp_physical) + 4L
)

wb_save(
  wb,
  file.path(results_dir, "fish_provisioning_results.xlsx"),
  overwrite = TRUE
)
```

## Joining results to the marine regions

Geometry lives in `data/gisdata/marine_regions.gpkg` and stays static. The published GeoPackage
holds one copy of each polygon, with one column per year for landings, resource rent and price.
Species, country and fleet belong in the RDS and Excel files, not
in the geometry. Duplicating the polygons once per year was what made the first export large;
three separate files, one per metric, would triplicate the geometry for no gain. QGIS users style
`tonnes_2023`, `rent_const_k_eur_2018` or `price_const_eur_kg_2023` on the same layer. Nominal rent
and price columns are kept beside the constant-price ones.

Two layers are published. The macro layer is the four pan-European sea regions; it joins on
`Sea_name` after a label recode. The division layer is the 58 FAO divisions and subdivisions; it
joins on `F_CODE`, or on `F_DIVISION` when that code is unique to one feature, which is how the
collapsed NEAO recodes find the dissolved polygons. Price is the quantity-weighted average,
revenue over live weight, not a sum of prices.

There is no legal-EEZ layer. `marine_regions.gpkg` is FAO fishing-area geometry only, and
`fish_provisioning.gpkg` therefore has the same two result layers. Methods Tables 10 to 13 label
FAO/GFCM `sub_reg_clean` as EEZ on the supply tables; that is the spatial unit of supply, not a
coastal-state exclusive economic zone. Landing-country tables use STECF `country_code` (Eq. 2) and
must not be drawn as if they were EEZs: the Member State of landing is not the coastal state of
the fishing ground. Maps stay on the FAO layers.

Mediterranean accounts use GFCM sub-areas (`sa1` to `sa27`) while the geometry is FAO `37.x.x`, so
the GSA lookup is applied before the join. Black Sea sub-areas `sa28` to `sa30` are not in that
workbook; they are mapped to Marmara, the Black Sea and the Azov Sea following the GFCM numbering.
ROW cannot appear on a European map. In 2023 it was 21 percent of tonnage, so any map needs that
reconciliation in mind. Legacy result columns on the source layers (`Tonnes_23`, `Money_23` and
the truncated overuse names) are not copied.

```{r}
#| label: write-gpkg

black_sea_to_fao <- tribble(
  ~sub_reg_clean, ~fao_division,
  "sa28", "37.4.1",
  "sa29", "37.4.2",
  "sa30", "37.4.3"
)

sub_reg_to_map_code <- bind_rows(
  gsa_to_fao |>
    transmute(
      sub_reg_clean = str_remove_all(sub_reg, "\\s"),
      map_code = fao_division
    ),
  black_sea_to_fao |>
    transmute(sub_reg_clean, map_code = fao_division)
)

sea_reg_to_name <- tribble(
  ~Sea_reg,             ~Sea_name,
  "NEAO",               "NEAO",
  "Mediterranean Sea",  "MED",
  "Black Sea",          "Black",
  "Baltic Sea",         "Baltic"
)

european_seas <- sea_reg_to_name$Sea_reg
panel_years <- sort(unique(fish_provisioning$year))

wide_metrics <- function(data, id_cols) {
  widen <- function(value_col, prefix) {
    data |>
      select(all_of(id_cols), year, all_of(value_col)) |>
      pivot_wider(
        names_from = year,
        values_from = all_of(value_col),
        names_prefix = prefix
      )
  }
  widen("landed_tonnes", "tonnes_") |>
    left_join(widen("revenue_thousand_eur", "rent_k_eur_"), by = id_cols) |>
    left_join(widen("revenue_thousand_eur_constant", "rent_const_k_eur_"), by = id_cols) |>
    left_join(widen("price_eur_per_kg", "price_eur_kg_"), by = id_cols) |>
    left_join(widen("price_eur_per_kg_constant", "price_const_eur_kg_"), by = id_cols) |>
    select(
      all_of(id_cols),
      all_of(paste0("tonnes_", panel_years)),
      all_of(paste0("rent_k_eur_", panel_years)),
      all_of(paste0("rent_const_k_eur_", panel_years)),
      all_of(paste0("price_eur_kg_", panel_years)),
      all_of(paste0("price_const_eur_kg_", panel_years))
    )
}

by_macro <- fish_provisioning |>
  filter(Sea_reg %in% european_seas) |>
  summarise(
    landed_kg = sum(landed_kg),
    revenue_eur = sum(revenue_eur),
    revenue_eur_constant = sum(revenue_eur_constant),
    .by = c(year, Sea_reg)
  ) |>
  left_join(sea_reg_to_name, by = "Sea_reg") |>
  mutate(
    landed_tonnes = landed_kg / 1000,
    revenue_thousand_eur = revenue_eur / 1000,
    revenue_thousand_eur_constant = revenue_eur_constant / 1000,
    price_eur_per_kg = if_else(landed_kg > 0, revenue_eur / landed_kg, NA_real_),
    price_eur_per_kg_constant = if_else(
      landed_kg > 0, revenue_eur_constant / landed_kg, NA_real_
    )
  )

macro_geom <- st_read(gis_file, layer = "sea_regions_macro", quiet = TRUE) |>
  select(Sea_name)

macro_out <- macro_geom |>
  left_join(wide_metrics(by_macro, c("Sea_name", "Sea_reg")), by = "Sea_name")

division_geom <- st_read(gis_file, layer = "sea_regions_division", quiet = TRUE) |>
  select(F_CODE, F_DIVISION, NAME_EN)

division_join_keys <- bind_rows(
  st_drop_geometry(division_geom) |>
    transmute(map_code = F_CODE, F_CODE),
  st_drop_geometry(division_geom) |>
    summarise(n_features = n(), F_CODE = F_CODE[1], .by = F_DIVISION) |>
    filter(n_features == 1) |>
    transmute(map_code = F_DIVISION, F_CODE)
) |>
  distinct(map_code, .keep_all = TRUE)

by_division <- fish_provisioning |>
  filter(Sea_reg %in% european_seas) |>
  left_join(sub_reg_to_map_code, by = "sub_reg_clean") |>
  mutate(map_code = coalesce(map_code, sub_reg_clean)) |>
  inner_join(division_join_keys, by = "map_code") |>
  summarise(
    landed_kg = sum(landed_kg),
    revenue_eur = sum(revenue_eur),
    revenue_eur_constant = sum(revenue_eur_constant),
    .by = c(year, F_CODE)
  ) |>
  mutate(
    landed_tonnes = landed_kg / 1000,
    revenue_thousand_eur = revenue_eur / 1000,
    revenue_thousand_eur_constant = revenue_eur_constant / 1000,
    price_eur_per_kg = if_else(landed_kg > 0, revenue_eur / landed_kg, NA_real_),
    price_eur_per_kg_constant = if_else(
      landed_kg > 0, revenue_eur_constant / landed_kg, NA_real_
    )
  )

division_out <- division_geom |>
  left_join(wide_metrics(by_division, "F_CODE"), by = "F_CODE")

gpkg_out <- file.path(gis_dir, "fish_provisioning.gpkg")
if (file.exists(gpkg_out)) {
  file.remove(gpkg_out)
}
st_write(macro_out, gpkg_out, layer = "sea_regions_macro", quiet = TRUE)
st_write(division_out, gpkg_out, layer = "sea_regions_division", quiet = TRUE)
```

A short coverage check for 2023: how much of the European-sea tonnage found a polygon, and how
much of the total was ROW and therefore off the map.

```{r}
#| label: check-map-coverage

mapped_2023 <- by_division |>
  filter(year == reference_year) |>
  summarise(mapped_tonnes = sum(landed_tonnes))

european_2023 <- fish_provisioning |>
  filter(year == reference_year, Sea_reg %in% european_seas) |>
  summarise(european_tonnes = sum(landed_tonnes))

row_2023 <- fish_provisioning |>
  filter(year == reference_year) |>
  summarise(
    total_tonnes = sum(landed_tonnes),
    row_tonnes = sum(landed_tonnes[Sea_reg == "ROW"])
  )

tibble(
  european_tonnes = european_2023$european_tonnes,
  mapped_tonnes = mapped_2023$mapped_tonnes,
  unmapped_european_tonnes = european_2023$european_tonnes - mapped_2023$mapped_tonnes,
  row_tonnes = row_2023$row_tonnes,
  row_share_pct = 100 * row_2023$row_tonnes / row_2023$total_tonnes
) |>
  gt() |>
  fmt_number(
    c(european_tonnes, mapped_tonnes, unmapped_european_tonnes, row_tonnes),
    decimals = 1
  ) |>
  fmt_number(row_share_pct, decimals = 1)
```

# Downloads

`_quarto.yml` lists `data/outputs/gis/` and `data/outputs/results/` under `project.resources`, so
these files are copied into the published site. The links below are relative to this page.

```{r}
#| label: downloads

tibble(
  File = c(
    "fish_provisioning.RDS",
    "fish_provisioning.xlsx",
    "fish_provisioning_results.xlsx",
    "rent_prices.RDS",
    "unvalued_records.RDS",
    "fish_provisioning.gpkg"
  ),
  Description = c(
    "Species-country-sub-region panel, 2008 to 2023, including 2010-euro rent",
    "The same panel as a single Excel sheet",
    "Report-style SUTs, one sheet per year: Europe Tables 2 to 9 and nested Tables 10 to 13 per sea, in tonnes, nominal euro and 2010 euro",
    "Observed resource-rent prices used in gap filling",
    "Landings that could not be valued",
    "Macro and division polygons with tonnes, nominal rent and 2010-euro rent"
  ),
  Path = c(
    file.path("../../../data/outputs/results", "fish_provisioning.RDS"),
    file.path("../../../data/outputs/results", "fish_provisioning.xlsx"),
    file.path("../../../data/outputs/results", "fish_provisioning_results.xlsx"),
    file.path("../../../data/outputs/results", "rent_prices.RDS"),
    file.path("../../../data/outputs/results", "unvalued_records.RDS"),
    file.path("../../../data/outputs/gis", "fish_provisioning.gpkg")
  )
) |>
  mutate(Download = sprintf("[%s](%s)", File, Path)) |>
  select(Download, Description) |>
  gt() |>
  fmt_markdown(Download) |>
  cols_label(Download = "File")
```

# Decisions from JRC (September 2026)

Domenico's answers to the open points, and what the routine now does.

- **Base year.** Constant prices use 2010 (`base_year` in the parameters chunk). Change that one
  line and re-render to switch; do not fold a new deflator into the allocation.
- **Subsidies on investments.** Operating subsidies (`totdirsub`) sit inside `variable_group ==
  "Income"`, so the Income plus Expenditure filter already captures them. Subsidies on investments
  (`subinvest`) sit under `Capital` and remain excluded. Box 1 of the methods report will be
  corrected so that it no longer appears to include them.
- **`Type` and `Type_1`.** The published tables nest species group (fish / shellfish / other) with
  fish type (benthic, benthopelagic, demersal, pelagic). That is our `species_group` and
  `fish_type`. In the classification workbook, `Type` is the water-column typology; the fish /
  shellfish split is the binary flags. There is no `Type_1` column. The original-script names were
  the other way around relative to Domenico's note; the table columns are what we match.
- **Duplicate species codes.** The eight duplicated codes are identical pairs. They are dropped
  with `distinct()`.
- **Mediterranean sub-areas `sa11.1` and `sa11.2`.** Recoded to `sa11` and counted in the
  Mediterranean. The 2023 shapefile tonne anchors for Mediterranean and ROW therefore shift by
  that tonnage.
- **EEZ.** STECF has no coastal-state EEZ field. Supply tables label `sub_reg_clean` as EEZ, with
  a note that these are FAO/GFCM fishing areas. Maps stay on FAO layers. Use tables are landing
  country.
- **Overuse.** Wired after allocation from the 2018 FSI scores. It is a companion overlay, not a
  column of the official SUT. Scores are held constant across years until a new stock-assessment
  vintage arrives.

# References

::: {#refs}
:::
