---
title: "Marine Ecosystem Services Processing"
subtitle: "Fish provisioning: annotated R processing routine"
format: html
toc: true
toc-depth: 3
bibliography: processing-references.bib
csl: ../../apa-6th-edition.csl
execute:
  warning: false
  message: false
---

# Introduction

This page documents the R routine that turns the Scientific, Technical and Economic Committee for
Fisheries (STECF) extraction [@stecf2024] into the fish provisioning accounts. Fish provisioning is
the ecosystem contribution to the growth of aquatic biomass captured in uncultivated production
contexts [@hainesyoung2018], and it is valued here with the resource rent approach of the SEEA
Central Framework [@un2014]. The companion page on
[methods](../marine-methods/index.qmd) states the equations; this page restates the earlier working
script in a compact, reproducible form so that the team can inspect each step, reproduce it from the
repository alone, and reuse the parts that generalise.

The routine reads two committed RDS files produced from the STECF fleet-segment workbook:
`FS_data_2008-2024.RDS` for income and expenditure, and `FS_landings_FAO_2008-2024.RDS` for
landings in kilograms and euro. Kilograms exist only in the landings file; income and expenditure
exist only in `FS data`. The conversion is documented on the
[STECF Fleet Segment Data](../../FS-landings-data/) page.

The routine has three components, and they run in this order because the second depends on the
first:

1. the **economic component**, which reduces the STECF economic variables to a single net economic
   result per fleet segment and year (Eq. 3 of the methods report);
2. the **biophysical component**, which cleans the landings table, harmonises fishing sub-regions,
   assigns each sub-region to a European sea region, and attaches the species typology;
3. the **resource rent allocation**, which distributes each fleet's net economic result across the
   sub-regions and species it landed, then fills the gaps left by fleets without usable economic
   data.

::: {.callout-important}
## 2024 is excluded from the valuation

The STECF files run to 2024, but no Member State has reported a usable expenditure series for that
year. The latest fully valuable year is 2023. The two year filters below therefore keep
`year <= reference_year` and drop 2024. Landings for 2024 can be shown in physical terms with a
provisional caveat; they cannot carry a resource rent.
:::

::: {.callout-note}
## Monetary values are 2008 euro

Allocation is done in nominal euro. A single step at the end converts resource rent to constant
2008 prices with Eurostat HICP for EU-27 [@eurostat2026hicp]. The base year is `base_year` in the
parameters chunk; changing it does not touch the allocation. Tables report both nominal and 2008
euro, plus the year-on-year real change in 2008 euro. Physical landings are not deflated.
:::

::: {.callout-note}
## Paths are relative to the project root

`_quarto.yml` sets `execute-dir: project`, so every path below resolves from the repository root
regardless of where the `.qmd` file sits. There is no need for `here::here()` or `setwd()`.
:::

## Preliminaries

We work in the tidyverse throughout. `readxl` reads the lookup workbooks, `sf` joins the results to
the marine region geometry, `writexl` writes the flat panel workbook, `openxlsx2` writes the
formatted results workbook in the Outputs section, and `gt` formats the inspection tables.
`writexl` cannot set number formats, column widths or borders, so it is kept only for the
unformatted panel.

```{r}
#| label: libraries

library(tidyverse)
library(readxl)
library(sf)
library(writexl)
library(openxlsx2)
library(gt)

sf_use_s2(FALSE)
```

Every input and output path is declared once, at the top, so that a reader can see the full data
footprint of the routine without scrolling. `reference_year` is the latest fully valuable year. The
two year filters use `year <= reference_year` rather than `year == reference_year`; `year` is
already part of every grouping key and every gap-filling join, so the result is a 2008 to 2023
panel with no cross-year contamination.

```{r}
#| label: parameters

reference_year <- 2023
base_year <- 2008

# Inputs
stecf_economic_file <- "data/stecf/FS_data_2008-2024.RDS"
stecf_landings_file <- "data/stecf/FS_landings_FAO_2008-2024.RDS"
country_codes_file <- "data/stecf/country_codes.csv"
species_lookup_file <- "data/lookup/3.classification_vs_OFFICIAL.xlsx"
gsa_lookup_file <- "data/lookup/Link_GSA_Marine regions_Mediterranean Sea.xlsx"
hicp_file <- "data/lookup/hicp_eu27.csv"
gis_file <- "data/gisdata/marine_regions.gpkg"

# Outputs
results_dir <- "data/outputs/results"
gis_dir <- "data/outputs/gis"
dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)
dir.create(gis_dir, showWarnings = FALSE, recursive = TRUE)

country_names <- readr::read_csv(country_codes_file, show_col_types = FALSE) |>
  transmute(country_code = `alpha-3`, country = name) |>
  mutate(
    country = recode(country, `Netherlands, Kingdom of the` = "Netherlands")
  )
```

# Reference tables

The classification decisions of the assessment were previously buried inside long `case_when()`
blocks or made by hand in Excel. They are collected here instead, as tables wherever a table will
do. The team can then review and amend them without reading the pipeline, and they are easy to
export for the annex of the report.

## Sub-region harmonisation

STECF reports fishing areas at a finer resolution than the accounts require, and it mixes two
naming systems: FAO fishing areas such as `37.1.1` and General Fisheries Commission for the
Mediterranean sub-areas such as `sa21`. The table below collapses the finer codes onto the code
actually used in the accounts. It reproduces the recodes of the original script exactly.

```{r}
#| label: sub-region-recodes
#| eval: true

sub_reg_recodes <- tribble(
  ~from,       ~to,
  "37.1.1",    "sa5",
  "sa21.1",    "sa21",
  "sa21.2",    "sa21",
  "sa21.3",    "sa21",
  "37.2.2",    "sa21",
  "37.1.3",    "sa21",
  "37.3.1",    "sa22",
  "27.2.a.1",  "27.2.a",
  "27.2.a.2",  "27.2.a",
  "27.2.b.1",  "27.2.b",
  "27.2.b.2",  "27.2.b",
  "27.3.a.20", "27.3.a",
  "27.3.a.21", "27.3.a",
  "27.3.c.22", "27.3.c",
  "27.3.b.23", "27.3.b",
  "27.5.a.1",  "27.5.a",
  "27.5.a.2",  "27.5.a",
  "27.5.b.1",  "27.5.b",
  "27.5.b.2",  "27.5.b",
  "27.14.b.1", "27.14.b",
  "27.14.b.2", "27.14.b"
)
```

The correspondence between GFCM sub-areas and FAO Mediterranean divisions is held separately in
`data/lookup`. The pipeline does not consume it; it is the reference that justifies the
Mediterranean rows of the recode table, and it is the authority against which the sea region rules
below can be audited.

```{r}
#| label: gsa-lookup
#| eval: true

gsa_to_fao <- read_excel(gsa_lookup_file, sheet = "Regions division") |>
  rename(sub_reg = `Row Labels`, fao_division = Sea_Region)
```


## Sea region assignment

Each harmonised sub-region belongs to one of the four pan-European marine regions of the Marine
Strategy Framework Directive, or to the rest of the world when a European fleet lands a catch taken
elsewhere. The rules are a direct transcription of the original `case_when()`; wrapping them in a
function means the same classification can be reused by the overuse and blue carbon routines
without copying the list of codes again.

```{r}
#| label: fn-assign-sea-region
#| eval: true

neao_sub_regions <- c(
  "27.1.a",  "27.1.b",  "27.10.a", "27.10.b", "27.2.a",  "27.2.b",
  "27.3.a",  "27.3.c",  "27.3.b",  "27.4.a",  "27.4.b",  "27.4.c",
  "27.5.a",  "27.5.b",  "27.6.a",  "27.6.b",  "27.7.a",  "27.7.b",
  "27.7.c",  "27.7.d",  "27.7.e",  "27.7.f",  "27.7.g",  "27.7.h",
  "27.7.j",  "27.7.k",  "27.8.a",  "27.8.b",  "27.8.c",  "27.8.d",
  "27.8.e",  "27.9.a",  "27.9.b",  "27.12.a", "27.12.b", "27.12.c",
  "27.14.a", "27.14.b"
)

assign_sea_region <- function(sub_reg_clean) {
  case_when(
    sub_reg_clean == "0" ~ "unknown",
    sub_reg_clean %in% c("sa28", "sa29", "sa30") ~ "Black Sea",
    sub_reg_clean %in% paste0("sa", 1:27) ~ "Mediterranean Sea",
    str_detect(sub_reg_clean, fixed("27.3.d")) ~ "Baltic Sea",
    str_detect(sub_reg_clean, fixed("27.8")) |
      sub_reg_clean %in% neao_sub_regions ~ "NEAO",
    .default = "ROW"
  )
}
```

## Species typology

STECF identifies species by their FAO ASFIS code but says nothing about what kind of animal it is
or where in the water column it lives. The methods report fills that gap from FishBase and
SeaLifeBase [@froese2024; @palomares2024], and the result is the classification workbook already in
`data/lookup`. We derive two attributes from it: `species_group`, which separates fish from
shellfish and from species whose code could not be matched with confidence, and `fish_type`, the
aggregated water column typology of Table 1 of the methods report.

```{r}
#| label: species-typology
#| eval: true

species_types_raw <- read_excel(species_lookup_file, sheet = "Classification")

species_types <- species_types_raw |>
  transmute(
    species_code,
    species_name_lookup = species_name,
    fish_type = Type,
    species_group = case_when(
      fish == 1 ~ "fish",
      shellfish == 1 ~ "shellfish",
      .default = "other"
    )
  ) |>
  distinct(species_code, .keep_all = TRUE)
```

::: {.callout-warning}
## This is where the manual Excel merge came from

The original script joined landings to this classification by hand in Excel because the join in R
produced double counting. The cause is visible in the workbook: eight species codes appear twice in
the `Classification` sheet, so a plain `left_join()` duplicates every landing record for those
species. `distinct(species_code, .keep_all = TRUE)` keeps the first entry and removes the
duplication, which makes the merge reproducible in R. The check chunk below lists the affected
codes so that the team can decide which of the two entries is correct.
:::

```{r}
#| label: check-duplicate-species
#| eval: true

species_types_raw |>
  filter(n() > 1, .by = species_code) |>
  arrange(species_code) |>
  select(species_code, species_name, Type, fish, shellfish, Other) |>
  gt()
```

# Helper functions

These operations recur often enough to be worth naming. Everything else stays as an explicit
pipeline, because a named function that is called once hides more than it saves.

Both STECF sheets carry the same block of administrative columns that the assessment never uses.
`any_of()` rather than `all_of()` means the function works on both sheets even though they do not
carry exactly the same set.

```{r}
#| label: fn-drop-admin-columns

drop_stecf_admin_columns <- function(data) {
  data |>
    select(-any_of(c(
      "upload_date", "geo_indicator", "cluster_name", "fromtable",
      "framework", "template_name", "gear", "fishery", "activity"
    )))
}
```

Harmonising sub-regions is two steps: strip the stray whitespace that STECF sometimes carries, then
apply the recode table. We strip first so that a code entered as `27.2.a.1 ` still matches. The raw
`sub_reg` is kept alongside `sub_reg_clean` for traceability.

```{r}
#| label: fn-harmonise-sub-reg
#| eval: true

harmonise_sub_reg <- function(data, recodes = sub_reg_recodes) {
  data |>
    mutate(sub_reg_clean = str_remove_all(sub_reg, "\\s")) |>
    left_join(
      recodes,
      by = join_by(sub_reg_clean == from),
      relationship = "many-to-one"
    ) |>
    mutate(sub_reg_clean = coalesce(to, sub_reg_clean)) |>
    select(-to)
}
```

The gap filling described in Box 2 of the methods report is the same operation applied three times
at successively coarser levels of aggregation. Expressing it once as a function that takes the
grouping columns as an argument removes about sixty lines of copied code and guarantees that the
three passes stay consistent with each other.

```{r}
#| label: fn-fill-price-gaps

# Mean observed price of a species over an arbitrary set of grouping columns.
average_price <- function(reference, by) {
  reference |>
    summarise(price = mean(price, na.rm = TRUE), .by = all_of(by))
}

# Assign that mean price to records that have landings but no usable rent.
# Records that still find no reference price keep a price of zero and are
# passed on to the next, coarser pass.
fill_price_gaps <- function(gaps, reference, by, basis) {
  gaps |>
    select(-any_of(c("price", "sum_revenue", "price_basis"))) |>
    left_join(
      average_price(reference, by),
      by = by,
      relationship = "many-to-one"
    ) |>
    mutate(
      price = replace_na(price, 0),
      price_basis = if_else(price > 0, basis, NA_character_),
      sum_revenue = sum_landed * price
    )
}
```

# Economic component

## Loading the STECF economic table

The economic file reports one row per fleet segment, year and economic variable. Inactive fleets
are removed first because they land nothing and would otherwise distort the fleet averages. The
year filter keeps the panel through `reference_year` and drops 2024, which cannot be valued.

```{r}
#| label: economic-load

stecf_economic <- readRDS(stecf_economic_file) |>
  filter(fishing_tech != "INACTIVE", year <= reference_year) |>
  drop_stecf_admin_columns()
```

## Net economic result by fleet

The resource rent approach values the service as the economic surplus of the fleet, that is total
income less total costs (Eq. 3 of the methods report). STECF splits its economic variables into
groups; we keep only `Expenditure` and `Income`, sign the expenditure negatively so that both can
be summed in a single column, and aggregate to one net result per fleet segment and year.

Missing values become zero and zero rows are then dropped. That looks circular, and it is
deliberate: STECF uses a blank to mean "not reported" and the assessment treats a non-reported item
as absent rather than as a true zero, so both end up excluded from the sum.

```{r}
#| label: fn-net-economic-by-fleet

net_economic_by_fleet <- function(data) {
  data |>
    filter(variable_group %in% c("Expenditure", "Income")) |>
    mutate(value = replace_na(value, 0)) |>
    filter(value != 0) |>
    mutate(money = if_else(variable_group == "Expenditure", -value, value)) |>
    summarise(
      total_expenditure = sum(money[variable_group == "Expenditure"]),
      total_income = sum(money[variable_group == "Income"]),
      total_net = total_income + total_expenditure,
      .by = c(year, fs_name)
    )
}

fleet_net <- net_economic_by_fleet(stecf_economic)
```

An ecosystem service cannot have a negative flow, because that would describe a disservice. Box 2
of the methods report therefore splits the fleets in two: those with a positive net result carry the
valuation, and those with a negative or zero result are treated as missing and recovered later from
average prices.

```{r}
#| label: fleet-net-positive

fleet_net_positive <- fleet_net |>
  filter(total_net > 0)
```

# Biophysical component

## Loading and cleaning landings

The landings file reports quantities in several units. Only kilograms are relevant here, and rows
with a zero quantity carry no information. We rename `value` to `landed_kg` immediately, because
three different quantities are called `value` at various points of the original script and the name
stops carrying meaning. The same `year <= reference_year` bound used on the economic file keeps the
two sides of the join on the same years.

```{r}
#| label: landings-load

landings_raw <- readRDS(stecf_landings_file) |>
  filter(year <= reference_year) |>
  drop_stecf_admin_columns() |>
  filter(unit == "kg", value != 0) |>
  rename(landed_kg = value)
```

## Harmonising regions and attaching the typology

Sub-region harmonisation, sea region assignment and the species join all happen in one pipeline.
The species join is the step that was previously done by hand in Excel; `relationship =
"many-to-one"` makes `dplyr` fail loudly if the lookup ever regains duplicate codes, which is the
guard the manual workflow lacked.

```{r}
#| label: landings-clean

landings <- landings_raw |>
  harmonise_sub_reg() |>
  mutate(Sea_reg = assign_sea_region(sub_reg_clean)) |>
  left_join(
    species_types,
    by = join_by(species_code),
    relationship = "many-to-one"
  )
```

Three checks are worth running every time the extraction is refreshed. The first lists species codes
present in the landings but absent from the classification workbook; those rows would otherwise
reach the accounting tables with a missing fish type. The second compares the species name STECF
reports against the name the classification workbook holds for the same code, which is how a
mis-assigned code shows itself. The third lists sub-regions that fell through to `ROW`, the
catch-all category and therefore the place where a new or misspelled STECF code will hide.

```{r}
#| label: check-unmatched

landings |>
  filter(year == reference_year, is.na(fish_type)) |>
  distinct(species_code, species_name) |>
  arrange(species_code) |>
  gt()

landings |>
  filter(year == reference_year, species_name != species_name_lookup) |>
  distinct(species_code, species_name, species_name_lookup) |>
  arrange(species_code) |>
  gt()

landings |>
  filter(year == reference_year, Sea_reg == "ROW") |>
  summarise(landed_kg = sum(landed_kg), .by = c(sub_reg, sub_reg_clean)) |>
  arrange(desc(landed_kg)) |>
  gt()
```

A fourth check needs no STECF data at all, so it is the one to look at first. Every sub-area listed
in the GFCM correspondence workbook is Mediterranean by definition; any of them that the recode and
classification rules fail to place in the Mediterranean is a gap in those rules.

```{r}
#| label: check-mediterranean-coverage
#| eval: true

gsa_to_fao |>
  harmonise_sub_reg() |>
  mutate(assigned = assign_sea_region(sub_reg_clean)) |>
  filter(assigned != "Mediterranean Sea") |>
  gt()
```

::: {.callout-warning}
## Known edge case in the Mediterranean codes

That last check returns two rows. `assign_sea_region()` matches Mediterranean sub-areas against the
literal list `sa1` to `sa27`, so the decimal sub-areas `sa11.1` and `sa11.2` do not match and fall
through to `ROW`. Both appear in the GFCM correspondence workbook, which suggests they are genuine
Mediterranean areas. We have kept the original behaviour rather than silently changing published
numbers; the `ROW` check above shows how much landing is affected, and the team can then decide
whether to add the two codes to `sub_reg_recodes`.
:::

The 2023 landings, once harmonised and assigned to a sea region, must reproduce the tonnage the
previous team baked into the macro-region shapefile. Any refactor of `harmonise_sub_reg()` or
`assign_sea_region()` that moves these figures has broken the biophysical account.

```{r}
#| label: check-2023-tonnes-anchor

anchor_2023 <- tribble(
  ~Sea_reg,             ~anchor_tonnes,
  "NEAO",                     1925814,
  "Baltic Sea",                436672,
  "Mediterranean Sea",         308722,
  "Black Sea",                   9961,
  "ROW",                       713131
)

anchor_check <- landings |>
  filter(year == reference_year) |>
  summarise(landed_tonnes = sum(landed_kg) / 1000, .by = Sea_reg) |>
  left_join(anchor_2023, by = "Sea_reg") |>
  mutate(
    difference_tonnes = landed_tonnes - anchor_tonnes,
    match = abs(difference_tonnes) < 0.6
  ) |>
  arrange(Sea_reg)

anchor_check |>
  gt() |>
  fmt_number(c(landed_tonnes, anchor_tonnes, difference_tonnes), decimals = 1)

stopifnot(
  "2023 sea-region tonnes must match the published shapefile anchors" =
    nrow(anchor_check) == nrow(anchor_2023) && all(anchor_check$match)
)
```

# Resource rent allocation

## Matching landings to fleet economics

Landings and fleet economics meet on the fleet segment name and the year. The join is
many-to-one because a fleet has one net result but many landing records. Records that find no
positive net result are the gap set of Box 2; they keep their quantities and are valued later from
average prices.

```{r}
#| label: landings-matched

landings_matched <- landings |>
  left_join(
    fleet_net_positive,
    by = join_by(fs_name, year),
    relationship = "many-to-one"
  )

landings_valued_input <- landings_matched |>
  filter(!is.na(total_net))

landings_without_rent <- landings_matched |>
  filter(is.na(total_net))
```

## Distributing the net result

A fleet earns one net result but fishes several sub-regions and several species, and STECF cannot
attribute cost or revenue any more finely than the fleet. The assessment therefore distributes the
net result in proportion to the quantities landed, in two nested steps: first across the
sub-regions the fleet worked (Eq. 4 and Eq. 5), then across the species it landed in each of those
sub-regions (Eq. 6 and Eq. 7).

The original script carried these weights as percentages and divided by one hundred again at each
use. We keep them as shares, which is arithmetically identical and removes four opportunities for a
misplaced factor of one hundred.

```{r}
#| label: rent-allocation

landings_valued <- landings_valued_input |>
  mutate(
    fleet_landed_kg = sum(landed_kg, na.rm = TRUE),
    .by = c(year, fs_name)
  ) |>
  mutate(
    fleet_area_landed_kg = sum(landed_kg, na.rm = TRUE),
    .by = c(year, fs_name, sub_reg_clean)
  ) |>
  mutate(
    # Eq. 4: share of the fleet's catch taken in this sub-region
    share_fleet_area = fleet_area_landed_kg / fleet_landed_kg,
    # Eq. 5: net result attributable to the fleet in this sub-region
    net_fleet_area = total_net * share_fleet_area,
    # Eq. 6: share of that sub-regional catch made up by this species
    share_species_area = landed_kg / fleet_area_landed_kg,
    # Eq. 7: net result attributable to the species in this sub-region
    net_species = net_fleet_area * share_species_area
  )
```

## Observed prices

The accounts report one figure per species, landing country and sub-region, so the fleet dimension
is aggregated away here. Dividing the summed net result by the summed quantity gives the implied
resource rent per kilogram, which is the price used to value the gap records in the next section.

```{r}
#| label: rent-prices

accounting_keys <- c(
  "year", "country_code", "sub_reg_clean", "Sea_reg",
  "species_code", "species_name", "species_group", "fish_type"
)

rent_prices <- landings_valued |>
  summarise(
    sum_landed = sum(landed_kg, na.rm = TRUE),
    sum_revenue = sum(net_species, na.rm = TRUE),
    .by = all_of(accounting_keys)
  ) |>
  mutate(
    price = if_else(sum_landed > 0, sum_revenue / sum_landed, NA_real_),
    price_basis = "observed rent"
  )
```

::: {.callout-note}
## Two deliberate differences from the original script

The original produced this table with `mutate()` followed by `distinct()`, which left one row per
underlying fleet record rather than one row per accounting key; because the raw `sub_reg` was among
the columns kept, sub-regions that had been collapsed by the recode table survived as separate rows
carrying the same group total, and the final aggregation added them together. Using `summarise()`
over an explicit key list produces exactly one row per group and removes that double counting.

The original also guarded the division against a zero in the individual record's own net value
rather than in the group's total quantity, which set a price to `NA` for records that had a valid
group price. The guard here is on `sum_landed`, which is the quantity actually in the denominator.
:::

## Gap set

Landings from fleets without a positive net result are aggregated onto the same accounting keys so
that the two halves of the dataset share one grain. They arrive with quantities and no revenue.

```{r}
#| label: price-gaps

price_gaps <- landings_without_rent |>
  summarise(
    sum_landed = sum(landed_kg, na.rm = TRUE),
    .by = all_of(accounting_keys)
  )
```

## Filling the gaps

Box 2 of the methods report fills the gaps in three passes, each one coarser than the last: the
average price of the species in the same country, then in the same sub-region, then across Europe.
Each pass keeps what it could value and hands the remainder to the next. Anything still unvalued
after the third pass has no observed price anywhere in the dataset for that species.

```{r}
#| label: gap-filling

# Pass 1: average price of the species in the same landing country
filled_country <- fill_price_gaps(
  price_gaps,
  rent_prices,
  by = c("year", "species_code", "species_name", "country_code"),
  basis = "country average"
)

# Pass 2: average price of the species in the same sub-region
filled_sub_reg <- filled_country |>
  filter(price == 0) |>
  fill_price_gaps(
    rent_prices,
    by = c("year", "species_code", "species_name", "sub_reg_clean"),
    basis = "sub-region average"
  )

# Pass 3: average price of the species anywhere in the dataset
filled_europe <- filled_sub_reg |>
  filter(price == 0) |>
  fill_price_gaps(
    rent_prices,
    by = c("year", "species_code", "species_name"),
    basis = "European average"
  )
```

The passes overlap by construction, so the first two contribute only the records they actually
valued. The third contributes everything that reaches it, including the records it could not value;
those carry a quantity and a revenue of zero, which is what keeps the physical account complete
even where the economic data does not reach. They are also extracted on their own, because their
size is a useful measure of that reach.

```{r}
#| label: gap-filling-bind

accounting_records <- bind_rows(
  rent_prices,
  filled_country |> filter(price > 0),
  filled_sub_reg |> filter(price > 0),
  filled_europe |> mutate(price_basis = replace_na(price_basis, "unvalued"))
)

unvalued_records <- filled_europe |>
  filter(price == 0)
```

# Final dataset

The four sources are aggregated once more onto the accounting keys, because a species, country and
sub-region combination can receive quantities from both a positive-rent fleet and a gap-filled
fleet. The final price is the revenue-weighted average that results, and the report's units are
tonnes and thousands of euro.

`price_basis` is collapsed into a readable label rather than dropped, so each row of the published
table carries the provenance of its own valuation. Note that both unit conversions are applied to
the aggregated columns; the original script converted quantities from the aggregate but converted
revenue from the pre-aggregation column, so the two published units were computed on different
bases.

```{r}
#| label: fish-provisioning

fish_provisioning <- accounting_records |>
  summarise(
    landed_kg = sum(sum_landed, na.rm = TRUE),
    revenue_eur = sum(sum_revenue, na.rm = TRUE),
    price_basis = str_c(sort(unique(price_basis)), collapse = " + "),
    .by = all_of(accounting_keys)
  ) |>
  mutate(
    price_eur_per_kg = if_else(
      landed_kg > 0, revenue_eur / landed_kg, NA_real_
    ),
    landed_tonnes = landed_kg / 1000,
    revenue_thousand_eur = revenue_eur / 1000
  ) |>
  arrange(year, Sea_reg, country_code, species_code)
```

Deflation is applied here, after the accounts exist, and nowhere else. Eurostat HICP for EU-27,
all-items, annual average index, is read from a committed lookup so the site builds without a
network call [@eurostat2026hicp]. The factor is `HICP_t / HICP_2008`. Resource rent in 2008 euro
is nominal rent divided by that factor. Year-on-year real change is computed later on the
aggregate tables, where each sea region has a complete annual series.

```{r}
#| label: deflate

hicp <- read_csv(hicp_file, show_col_types = FALSE)
hicp_base <- hicp$hicp_index[hicp$year == base_year]
if (length(hicp_base) != 1 || is.na(hicp_base) || hicp_base <= 0) {
  stop("HICP lookup has no usable index for base_year = ", base_year)
}

fish_provisioning <- fish_provisioning |>
  left_join(hicp, by = "year") |>
  mutate(
    deflator = hicp_index / hicp_base,
    revenue_eur_2008 = revenue_eur / deflator,
    revenue_thousand_eur_2008 = revenue_eur_2008 / 1000,
    price_eur_per_kg_2008 = if_else(
      landed_kg > 0, revenue_eur_2008 / landed_kg, NA_real_
    )
  )

hicp |>
  mutate(
    deflator = hicp_index / hicp_base,
    base = if_else(year == base_year, "base", "")
  ) |>
  gt() |>
  cols_label(
    year = "Year",
    hicp_index = "HICP (2015 = 100)",
    deflator = paste0("Deflator (", base_year, " = 1)"),
    base = ""
  ) |>
  fmt_number(hicp_index, decimals = 2) |>
  fmt_number(deflator, decimals = 4) |>
  tab_source_note(
    "Eurostat prc_hicp_aind, EU-27 (from 2020), CP00, annual average index."
  )
```

The headline supply figure of the report is the catch by sea region, in tonnes and in thousands of
euro. That is Eq. 1. STECF `country_code` is the Member State where the catch was landed, which is
the use-table grain of Eq. 2. `fish_type` is the water-column typology of methods Table 1 (`Type` in
the original script). The published annexes also split supply by coastal-state EEZ (Tables 10 to 13);
that geometry is not in the repository, so EEZ tables and maps are not produced here. The two tables
below are the 2023 slice, in 2008 euro.

```{r}
#| label: supply-summary

fish_provisioning |>
  filter(year == reference_year) |>
  summarise(
    landed_tonnes = sum(landed_tonnes),
    revenue_thousand_eur = sum(revenue_thousand_eur),
    revenue_thousand_eur_2008 = sum(revenue_thousand_eur_2008),
    .by = Sea_reg
  ) |>
  arrange(desc(landed_tonnes)) |>
  gt() |>
  fmt_number(decimals = 2) |>
  tab_header(title = "Eq. 1. Supply by sea region, 2023") |>
  tab_source_note(
    paste0("Nominal euro and constant ", base_year, " euro. Latest fully valuable year.")
  )

fish_provisioning |>
  filter(year == reference_year) |>
  left_join(country_names, by = "country_code") |>
  mutate(country = coalesce(country, country_code)) |>
  summarise(
    landed_tonnes = sum(landed_tonnes),
    revenue_thousand_eur = sum(revenue_thousand_eur),
    revenue_thousand_eur_2008 = sum(revenue_thousand_eur_2008),
    .by = c(Sea_reg, country)
  ) |>
  arrange(Sea_reg, country) |>
  gt() |>
  fmt_number(decimals = 2) |>
  tab_header(title = "Eq. 2. Use by sea region and landing country, 2023") |>
  cols_label(Sea_reg = "Sea region", country = "Country") |>
  tab_source_note(
    "Country is the Member State of landing, not the EEZ of the fishing ground."
  )
```

# Outputs

The original script wrote twelve intermediate workbooks to disk. We write the panel twice as a
flat deliverable, once as RDS and once as Excel, so the client can open the same table in either
tool. The aggregate use tables, one year per sheet, are a third workbook. Column A is sea region by
landing country by fish type, with missing fish types filled by zero; column K is the shorter sea
region by landing country table. Observed prices and unvalued catch stay as RDS.

```{r}
#| label: write-outputs

saveRDS(fish_provisioning, file.path(results_dir, "fish_provisioning.RDS"))
saveRDS(rent_prices, file.path(results_dir, "rent_prices.RDS"))
saveRDS(unvalued_records, file.path(results_dir, "unvalued_records.RDS"))

write_xlsx(
  list(fish_provisioning = fish_provisioning),
  path = file.path(results_dir, "fish_provisioning.xlsx")
)

sea_order <- c(
  "NEAO", "Baltic Sea", "Mediterranean Sea", "Black Sea", "ROW"
)
fish_type_order <- c(
  "benthic", "benthopelagic", "demersal", "pelagic", "unclassified"
)

supply_detail <- fish_provisioning |>
  filter(Sea_reg != "unknown") |>
  mutate(
    fish_type = coalesce(fish_type, "unclassified"),
    Sea_reg = factor(Sea_reg, levels = sea_order),
    fish_type = factor(fish_type, levels = fish_type_order)
  ) |>
  summarise(
    landed_tonnes = sum(landed_tonnes),
    revenue_thousand_eur = sum(revenue_thousand_eur),
    revenue_thousand_eur_2008 = sum(revenue_thousand_eur_2008),
    .by = c(year, Sea_reg, country_code, fish_type)
  ) |>
  complete(
    nesting(year, Sea_reg, country_code),
    fish_type,
    fill = list(
      landed_tonnes = 0,
      revenue_thousand_eur = 0,
      revenue_thousand_eur_2008 = 0
    )
  ) |>
  left_join(country_names, by = "country_code") |>
  mutate(
    country = coalesce(country, country_code),
    price_eur_per_kg_2008 = if_else(
      landed_tonnes > 0,
      revenue_thousand_eur_2008 / landed_tonnes,
      NA_real_
    )
  ) |>
  arrange(Sea_reg, country, fish_type, year) |>
  mutate(
    real_change_pct = if_else(
      lag(revenue_thousand_eur_2008) > 0,
      100 * (revenue_thousand_eur_2008 / lag(revenue_thousand_eur_2008) - 1),
      NA_real_
    ),
    .by = c(Sea_reg, country_code, fish_type)
  )

supply_compact <- supply_detail |>
  summarise(
    landed_tonnes = sum(landed_tonnes),
    revenue_thousand_eur = sum(revenue_thousand_eur),
    revenue_thousand_eur_2008 = sum(revenue_thousand_eur_2008),
    .by = c(year, Sea_reg, country_code, country)
  ) |>
  mutate(
    price_eur_per_kg_2008 = if_else(
      landed_tonnes > 0,
      revenue_thousand_eur_2008 / landed_tonnes,
      NA_real_
    )
  ) |>
  arrange(Sea_reg, country, year) |>
  mutate(
    real_change_pct = if_else(
      lag(revenue_thousand_eur_2008) > 0,
      100 * (revenue_thousand_eur_2008 / lag(revenue_thousand_eur_2008) - 1),
      NA_real_
    ),
    .by = c(Sea_reg, country_code)
  )

supply_totals <- supply_compact |>
  summarise(
    landed_tonnes = sum(landed_tonnes),
    revenue_thousand_eur = sum(revenue_thousand_eur),
    revenue_thousand_eur_2008 = sum(revenue_thousand_eur_2008),
    .by = year
  ) |>
  arrange(year) |>
  mutate(
    Sea_reg = "Total",
    country = NA_character_,
    fish_type = NA_character_,
    price_eur_per_kg_2008 = if_else(
      landed_tonnes > 0,
      revenue_thousand_eur_2008 / landed_tonnes,
      NA_real_
    ),
    real_change_pct = if_else(
      lag(revenue_thousand_eur_2008) > 0,
      100 * (revenue_thousand_eur_2008 / lag(revenue_thousand_eur_2008) - 1),
      NA_real_
    )
  )

sheet_detail <- function(y) {
  bind_rows(
    supply_detail |>
      filter(year == y) |>
      arrange(Sea_reg, country, fish_type) |>
      mutate(
        Sea_reg = as.character(Sea_reg),
        fish_type = as.character(fish_type)
      ),
    supply_totals |> filter(year == y)
  ) |>
    transmute(
      sea_region = Sea_reg,
      country,
      fish_type,
      landed_tonnes,
      revenue_thousand_eur,
      revenue_thousand_eur_2008,
      price_eur_per_kg_2008,
      real_change_pct
    )
}

sheet_summary <- function(y) {
  bind_rows(
    supply_compact |>
      filter(year == y) |>
      arrange(Sea_reg, country) |>
      mutate(Sea_reg = as.character(Sea_reg)),
    supply_totals |> filter(year == y)
  ) |>
    transmute(
      sea_region = Sea_reg,
      country,
      landed_tonnes,
      revenue_thousand_eur,
      revenue_thousand_eur_2008,
      price_eur_per_kg_2008,
      real_change_pct
    )
}

result_years <- sort(unique(fish_provisioning$year), decreasing = TRUE)

detail_headers <- c(
  "Sea region",
  "Country",
  "Fish type",
  "Landings (tonnes)",
  "Resource rent (thousand nominal euro)",
  "Resource rent (thousand 2008 euro)",
  "Price (2008 euro per kg)",
  "Real change (percent)"
)
summary_headers <- c(
  "Sea region",
  "Country",
  "Landings (tonnes)",
  "Resource rent (thousand nominal euro)",
  "Resource rent (thousand 2008 euro)",
  "Price (2008 euro per kg)",
  "Real change (percent)"
)
header_row <- 5L
detail_start <- 1L
summary_start <- 11L
col_width <- 10
spacer_width <- 3
number_format <- "#,##0.00"

write_banner <- function(wb, sheet, year, start_col, n_cols, grain) {
  end_col <- start_col + n_cols - 1L
  lines <- c(
    "European Commission, Joint Research Centre (JRC)",
    "Fish provisioning ecosystem service accounts",
    paste0(
      "Year ", year, ". ", grain,
      " Resource rent in thousand euro. Constant prices use ",
      base_year, " as the base year (Eurostat HICP, EU-27)."
    )
  )
  for (row in 1:3) {
    dims <- wb_dims(rows = row, cols = start_col:end_col)
    wb <- wb_add_data(
      wb, sheet, x = lines[[row]],
      start_row = row, start_col = start_col, col_names = FALSE
    )
    wb <- wb_merge_cells(wb, sheet, dims = dims)
    wb <- wb_add_cell_style(
      wb, sheet,
      dims = wb_dims(rows = row, cols = start_col),
      wrap_text = TRUE, vertical = "center"
    )
  }
  wb_set_row_heights(wb, sheet, rows = 1:3, heights = 18)
}

write_table <- function(wb, sheet, data, start_col, headers) {
  n_cols <- length(headers)
  end_col <- start_col + n_cols - 1L
  header_dims <- wb_dims(rows = header_row, cols = start_col:end_col)
  first_numeric <- n_cols - 4L
  data_start <- header_row + 1L
  data_end <- header_row + nrow(data)

  wb <- wb_add_data(
    wb, sheet, x = matrix(headers, nrow = 1),
    start_row = header_row, start_col = start_col, col_names = FALSE
  )
  wb <- wb_add_font(wb, sheet, dims = header_dims, bold = TRUE)
  wb <- wb_add_cell_style(
    wb, sheet, dims = header_dims,
    wrap_text = TRUE, horizontal = "center", vertical = "center"
  )
  wb <- wb_add_border(
    wb, sheet, dims = header_dims,
    bottom_border = "thin",
    top_border = "none",
    left_border = "none",
    right_border = "none"
  )
  wb <- wb_set_row_heights(wb, sheet, rows = header_row, heights = 60)
  wb <- wb_add_data(
    wb, sheet, x = data,
    start_row = data_start, start_col = start_col,
    col_names = FALSE, na = "", apply_cell_style = FALSE
  )
  wb <- wb_add_numfmt(
    wb, sheet,
    dims = wb_dims(
      rows = data_start:data_end,
      cols = (start_col + first_numeric - 1L):end_col
    ),
    numfmt = number_format
  )
  for (row in which(data[[1]] == "Total") + header_row) {
    total_dims <- wb_dims(rows = row, cols = start_col:end_col)
    wb <- wb_add_font(wb, sheet, dims = total_dims, bold = TRUE)
    wb <- wb_add_border(
      wb, sheet, dims = total_dims,
      top_border = "thin",
      bottom_border = "thin",
      left_border = "none",
      right_border = "none"
    )
    wb <- wb_set_row_heights(wb, sheet, rows = row, heights = 18)
  }
  wb_set_col_widths(wb, sheet, cols = start_col:end_col, widths = col_width)
}

wb <- wb_workbook()
wb <- wb_add_worksheet(wb, "Readme")
readme_notes <- c(
  "European Commission, Joint Research Centre (JRC)",
  "Fish provisioning ecosystem service accounts",
  paste0(
    "Base year ", base_year,
    ". Resource rent in thousand euro. Constant prices use Eurostat HICP for EU-27."
  ),
  "Country is the STECF Member State where the catch was landed (Eq. 2), shown as the ISO name.",
  "Left table (column A): sea region by landing country by fish type. Missing fish types are filled with zero.",
  "Right table (column K): sea region by landing country, without fish type.",
  "Fish type is the water-column typology (Type in the original script). Unmatched codes are unclassified.",
  "EEZ supply tables (methods Tables 10 to 13) need coastal-state EEZ polygons, which are not in the repository.",
  "Real change is the year-on-year change in 2008 euro. Blank in the base year and when the previous value is zero.",
  "ROW is rest of world and is not shown on the European map.",
  "Source: STECF 2025 Fleet Economic data call."
)
wb <- wb_add_data(wb, "Readme", x = readme_notes, col_names = FALSE)
wb <- wb_add_cell_style(
  wb, "Readme",
  dims = wb_dims(rows = seq_along(readme_notes), cols = 1),
  wrap_text = TRUE
)
wb <- wb_set_col_widths(wb, "Readme", cols = 1, widths = 80)

for (y in result_years) {
  year_name <- as.character(y)
  wb <- wb_add_worksheet(wb, year_name)
  wb <- write_banner(
    wb, year_name, y, detail_start, length(detail_headers),
    "Sea region by landing country by fish type."
  )
  wb <- write_banner(
    wb, year_name, y, summary_start, length(summary_headers),
    "Sea region by landing country."
  )
  wb <- write_table(wb, year_name, sheet_detail(y), detail_start, detail_headers)
  wb <- write_table(wb, year_name, sheet_summary(y), summary_start, summary_headers)
  spacer_cols <- (detail_start + length(detail_headers)):(summary_start - 1L)
  wb <- wb_set_col_widths(wb, year_name, cols = spacer_cols, widths = spacer_width)
  wb <- wb_freeze_pane(
    wb, year_name,
    first_active_row = header_row + 1L, first_active_col = 1L
  )
}

wb_save(
  wb,
  file.path(results_dir, "fish_provisioning_results.xlsx"),
  overwrite = TRUE
)

```

## Joining results to the marine regions

Geometry lives in `data/gisdata/marine_regions.gpkg` and stays static. The published GeoPackage
holds one copy of each polygon, with one column per year for landings, resource rent and price.
Species, country and fleet belong in the RDS and Excel files, not
in the geometry. Duplicating the polygons once per year was what made the first export large;
three separate files, one per metric, would triplicate the geometry for no gain. QGIS users style
`tonnes_2023`, `rent08_k_eur_2018` or `price08_eur_kg_2023` on the same layer. Nominal rent and
price columns are kept beside the 2008-euro ones.

Two layers are published. The macro layer is the four pan-European sea regions; it joins on
`Sea_name` after a label recode. The division layer is the 58 FAO divisions and subdivisions; it
joins on `F_CODE`, or on `F_DIVISION` when that code is unique to one feature, which is how the
collapsed NEAO recodes find the dissolved polygons. Price is the quantity-weighted average,
revenue over live weight, not a sum of prices.

There is no EEZ layer. `marine_regions.gpkg` is FAO fishing-area geometry only, and
`fish_provisioning.gpkg` therefore has the same two result layers. Methods Tables 10 to 13 and
the annex maps split supply by coastal-state EEZ inside each sea region. That needs an EEZ
polygon layer, which was not delivered with the source shapefiles. Landing-country tables use
STECF `country_code` (Eq. 2) and must not be drawn as if they were EEZs: the Member State of
landing is not the coastal state of the fishing ground. Once EEZ geometry arrives, join a
year-column summary to it the same way as the macro layer; do not explode polygons by year.

Mediterranean accounts use GFCM sub-areas (`sa1` to `sa27`) while the geometry is FAO `37.x.x`, so
the GSA lookup is applied before the join. Black Sea sub-areas `sa28` to `sa30` are not in that
workbook; they are mapped to Marmara, the Black Sea and the Azov Sea following the GFCM numbering.
ROW cannot appear on a European map. In 2023 it was 21 percent of tonnage, so any map needs that
reconciliation in mind. Legacy result columns on the source layers (`Tonnes_23`, `Money_23` and
the truncated overuse names) are not copied.

```{r}
#| label: write-gpkg

black_sea_to_fao <- tribble(
  ~sub_reg_clean, ~fao_division,
  "sa28", "37.4.1",
  "sa29", "37.4.2",
  "sa30", "37.4.3"
)

sub_reg_to_map_code <- bind_rows(
  gsa_to_fao |>
    transmute(
      sub_reg_clean = str_remove_all(sub_reg, "\\s"),
      map_code = fao_division
    ),
  black_sea_to_fao |>
    transmute(sub_reg_clean, map_code = fao_division)
)

sea_reg_to_name <- tribble(
  ~Sea_reg,             ~Sea_name,
  "NEAO",               "NEAO",
  "Mediterranean Sea",  "MED",
  "Black Sea",          "Black",
  "Baltic Sea",         "Baltic"
)

european_seas <- sea_reg_to_name$Sea_reg
panel_years <- sort(unique(fish_provisioning$year))

wide_metrics <- function(data, id_cols) {
  widen <- function(value_col, prefix) {
    data |>
      select(all_of(id_cols), year, all_of(value_col)) |>
      pivot_wider(
        names_from = year,
        values_from = all_of(value_col),
        names_prefix = prefix
      )
  }
  widen("landed_tonnes", "tonnes_") |>
    left_join(widen("revenue_thousand_eur", "rent_k_eur_"), by = id_cols) |>
    left_join(widen("revenue_thousand_eur_2008", "rent08_k_eur_"), by = id_cols) |>
    left_join(widen("price_eur_per_kg", "price_eur_kg_"), by = id_cols) |>
    left_join(widen("price_eur_per_kg_2008", "price08_eur_kg_"), by = id_cols) |>
    select(
      all_of(id_cols),
      all_of(paste0("tonnes_", panel_years)),
      all_of(paste0("rent_k_eur_", panel_years)),
      all_of(paste0("rent08_k_eur_", panel_years)),
      all_of(paste0("price_eur_kg_", panel_years)),
      all_of(paste0("price08_eur_kg_", panel_years))
    )
}

by_macro <- fish_provisioning |>
  filter(Sea_reg %in% european_seas) |>
  summarise(
    landed_kg = sum(landed_kg),
    revenue_eur = sum(revenue_eur),
    revenue_eur_2008 = sum(revenue_eur_2008),
    .by = c(year, Sea_reg)
  ) |>
  left_join(sea_reg_to_name, by = "Sea_reg") |>
  mutate(
    landed_tonnes = landed_kg / 1000,
    revenue_thousand_eur = revenue_eur / 1000,
    revenue_thousand_eur_2008 = revenue_eur_2008 / 1000,
    price_eur_per_kg = if_else(landed_kg > 0, revenue_eur / landed_kg, NA_real_),
    price_eur_per_kg_2008 = if_else(
      landed_kg > 0, revenue_eur_2008 / landed_kg, NA_real_
    )
  )

macro_geom <- st_read(gis_file, layer = "sea_regions_macro", quiet = TRUE) |>
  select(Sea_name)

macro_out <- macro_geom |>
  left_join(wide_metrics(by_macro, c("Sea_name", "Sea_reg")), by = "Sea_name")

division_geom <- st_read(gis_file, layer = "sea_regions_division", quiet = TRUE) |>
  select(F_CODE, F_DIVISION, NAME_EN)

division_join_keys <- bind_rows(
  st_drop_geometry(division_geom) |>
    transmute(map_code = F_CODE, F_CODE),
  st_drop_geometry(division_geom) |>
    summarise(n_features = n(), F_CODE = F_CODE[1], .by = F_DIVISION) |>
    filter(n_features == 1) |>
    transmute(map_code = F_DIVISION, F_CODE)
) |>
  distinct(map_code, .keep_all = TRUE)

by_division <- fish_provisioning |>
  filter(Sea_reg %in% european_seas) |>
  left_join(sub_reg_to_map_code, by = "sub_reg_clean") |>
  mutate(map_code = coalesce(map_code, sub_reg_clean)) |>
  inner_join(division_join_keys, by = "map_code") |>
  summarise(
    landed_kg = sum(landed_kg),
    revenue_eur = sum(revenue_eur),
    revenue_eur_2008 = sum(revenue_eur_2008),
    .by = c(year, F_CODE)
  ) |>
  mutate(
    landed_tonnes = landed_kg / 1000,
    revenue_thousand_eur = revenue_eur / 1000,
    revenue_thousand_eur_2008 = revenue_eur_2008 / 1000,
    price_eur_per_kg = if_else(landed_kg > 0, revenue_eur / landed_kg, NA_real_),
    price_eur_per_kg_2008 = if_else(
      landed_kg > 0, revenue_eur_2008 / landed_kg, NA_real_
    )
  )

division_out <- division_geom |>
  left_join(wide_metrics(by_division, "F_CODE"), by = "F_CODE")

gpkg_out <- file.path(gis_dir, "fish_provisioning.gpkg")
if (file.exists(gpkg_out)) {
  file.remove(gpkg_out)
}
st_write(macro_out, gpkg_out, layer = "sea_regions_macro", quiet = TRUE)
st_write(division_out, gpkg_out, layer = "sea_regions_division", quiet = TRUE)
```

A short coverage check for 2023: how much of the European-sea tonnage found a polygon, and how
much of the total was ROW and therefore off the map.

```{r}
#| label: check-map-coverage

mapped_2023 <- by_division |>
  filter(year == reference_year) |>
  summarise(mapped_tonnes = sum(landed_tonnes))

european_2023 <- fish_provisioning |>
  filter(year == reference_year, Sea_reg %in% european_seas) |>
  summarise(european_tonnes = sum(landed_tonnes))

row_2023 <- fish_provisioning |>
  filter(year == reference_year) |>
  summarise(
    total_tonnes = sum(landed_tonnes),
    row_tonnes = sum(landed_tonnes[Sea_reg == "ROW"])
  )

tibble(
  european_tonnes = european_2023$european_tonnes,
  mapped_tonnes = mapped_2023$mapped_tonnes,
  unmapped_european_tonnes = european_2023$european_tonnes - mapped_2023$mapped_tonnes,
  row_tonnes = row_2023$row_tonnes,
  row_share_pct = 100 * row_2023$row_tonnes / row_2023$total_tonnes
) |>
  gt() |>
  fmt_number(
    c(european_tonnes, mapped_tonnes, unmapped_european_tonnes, row_tonnes),
    decimals = 1
  ) |>
  fmt_number(row_share_pct, decimals = 1)
```

# Downloads

`_quarto.yml` lists `data/outputs/gis/` and `data/outputs/results/` under `project.resources`, so
these files are copied into the published site. The links below are relative to this page.

```{r}
#| label: downloads

tibble(
  File = c(
    "fish_provisioning.RDS",
    "fish_provisioning.xlsx",
    "fish_provisioning_results.xlsx",
    "rent_prices.RDS",
    "unvalued_records.RDS",
    "fish_provisioning.gpkg"
  ),
  Description = c(
    "Species-country-sub-region panel, 2008 to 2023, including 2008-euro rent",
    "The same panel as a single Excel sheet",
    "Two tables per year: sea region by country by fish type (column A) and by country (column K)",
    "Observed resource-rent prices used in gap filling",
    "Landings that could not be valued",
    "Macro and division polygons with tonnes, nominal rent and 2008-euro rent"
  ),
  Path = c(
    file.path("../../../data/outputs/results", "fish_provisioning.RDS"),
    file.path("../../../data/outputs/results", "fish_provisioning.xlsx"),
    file.path("../../../data/outputs/results", "fish_provisioning_results.xlsx"),
    file.path("../../../data/outputs/results", "rent_prices.RDS"),
    file.path("../../../data/outputs/results", "unvalued_records.RDS"),
    file.path("../../../data/outputs/gis", "fish_provisioning.gpkg")
  )
) |>
  mutate(Download = sprintf("[%s](%s)", File, Path)) |>
  select(Download, Description) |>
  gt() |>
  fmt_markdown(Download) |>
  cols_label(Download = "File")
```

# Open points

These are decisions carried over from the original script, or deferred so that a later choice does
not force a rewrite.

- **Base year.** Monetary values are currently 2008 euro. The Annual Economic Report uses 2023
  constant prices. If JRC prefers another base, change `base_year` in the parameters chunk
  and re-render; do not fold a new deflator into the allocation.
- **Subsidies on investments.** Operating subsidies (`totdirsub`) sit inside `variable_group ==
  "Income"`, so the Income plus Expenditure filter already captures them. Subsidies on investments
  (`subinvest`) sit under `Capital` and are excluded. Box 1 of the methods report arguably includes
  them.
- **`Type` and `Type_1`.** The original script carried two classification columns under these
  names. We have read `Type` as the water column typology and `Type_1` as the fish, shellfish and
  other split, and renamed them `fish_type` and `species_group`. This should be confirmed against
  the published accounting tables.
- **Duplicate species codes.** Eight codes appear twice in the classification workbook. The routine
  keeps the first entry; the correct entry should be chosen and the workbook corrected at source.
- **Mediterranean sub-areas `sa11.1` and `sa11.2`.** See the warning above; these currently fall
  into `ROW`.
- **EEZ geometry.** Coastal-state EEZ polygons were not in the source GIS bundle. Eq. 2 country
  tables use landing Member State. EEZ supply tables and maps wait on that geometry.
- **Overuse / sustainability score.** Not wired. The lookup is `data/lookup/Overuse_Total.xlsx`;
  the method is chapter 4.5 of the methods report and `reference/AP_Deliverable4.pdf`. Apply it
  after allocation, as a last step like HICP, not inside Eq. 3 to 7.

# References

::: {#refs}
:::
